Proteostasis modulates gene dosage evolution in antibiotic-resistant bacteria

Version 1

Abstract

Abstract

Evolution of gene expression frequently drives antibiotic resistance in bacteria. We had previously (Patel and Matange, eLife, 2021) shown that in Escherichia coli, mutations at the mgrB locus were beneficial in trimethoprim and led to overexpression of dihydrofolate reductase (DHFR), encoded by the folA gene. Here, we show that DHFR levels are further enhanced by spontaneous duplication of a genomic segment encompassing folA and spanning hundreds of kilobases. This duplication was rare in wild type E. coli. However, its frequency was elevated in a lon-knockout strain, altering the mutational landscape early during trimethoprim adaptation. We then exploit this system to investigate the relationship between trimethoprim pressure and folA copy number. During long-term evolution, folA duplications were frequently reversed. Reversal was slower under antibiotic pressure, first requiring the acquisition of point mutations in DHFR or its promoter. Unexpectedly, despite resistance-conferring point mutations, some populations under high trimethoprim pressure maintained folA duplication to compensate for low abundance DHFR mutants. We find that evolution of gene dosage depends on expression demand, which is generated by antibiotic and exacerbated by proteolysis of drug-resistant mutants of DHFR. We propose a novel role for proteostasis as a determinant of copy number evolution in antibiotic-resistant bacteria.

Introduction

The dosage of a gene, i.e., copy number relative to the genome, directly impacts its level of expression and can alter organismal phenotype and fitness [1–3]. Mutations that increase copy numbers, termed as gene duplications and amplifications (GDAs), are implicated in diverse biological phenomena including cancer progression [4–6], speciation [7–9], reproductive isolation [10–12] and antimicrobial resistance (AMR) [2, 13]. GDAs are important for rapid adaptation since their frequencies of occurrence are higher by several orders of magnitude (∼1000-fold) than point mutations [2, 13–16]. Following the initial GDA mutation, however, amplified genes could be lost over time, pseudogenized, sub-functionalized or neo-functionalized [17–19]. Which of these possibilities is realized depends on the nature and strength of selection imposed by the environment, as well as the fitness cost associated with maintaining multiple gene copies [17–19].

In bacteria, GDA mutations are frequently encountered during adaptation to antibiotics [13] or heavy metals [20, 21]. They occur by RecA-dependent nonequal recombination as well as RecA-independent processes and often involve repeat sequences such as in IS-elements [22–25]. Amplified genes in drug-resistant bacteria code for effectors of resistance such as drug inactivating enzymes, drug target modifiers or “by-pass” mechanisms [26–28], but may also compensate for resistance-associated fitness costs [29]. More recently, the view that GDAs are less “hard-wired” than point mutations has emerged. For instance, transient alterations in gene dosage are a common mechanism of heteroresistance, i.e. resistant sub-populations within a drug-sensitive bacterial isolate [20, 30, 31]. There is also evidence to suggest that copy number variations are more common in bacteria than appreciated. It is estimated that as many as 20% of cells in a bacterial population may harbour a duplication of some genomic region [32]. These copy number polymorphisms create within-population heterogeneities in gene expression, serving as a bet-hedging strategy with adaptive value in fluctuating environments [33].

Dihydrofolate reductase (DHFR) is a ubiquitous enzyme crucial for growth and division. It catalyzes the formation of tetrahydrofolate, an important cofactor in nucleotide and amino acid biosynthesis. DHFR is coded by an essential gene in most organisms and its deletion or pharmacological inhibition results in cell death or growth stasis [34]. Not surprisingly, anti-folates are effective therapeutic agents for diverse pathologies [35, 36]. For instance, methotrexate, used in the treatment of cancers and autoimmune conditions [37], pyrimethamine, used as an anti-malarial [38], and trimethoprim, used as an anti-bacterial [39], all of which are competitive inhibitors of DHFR. Interestingly, copy number amplification and consequent overexpression of DHFR confers resistance in all three systems [40–43]. Thus, the gene dosage of DHFR appears to be tightly regulated across evolutionarily divergent organisms, making it a useful system to understand the principles underlying selection and evolutionary fate of GDA mutations.

In Escherichia coli DHFR is coded by the folA gene. Several mutation hot spots for trimethoprim resistance have been identified in this enzyme [44–47]. Additionally, mechanisms that transcriptionally up-regulate folA expression are also known in trimethoprim resistant E. coli. These include cis-regulatory mutations in the folA promoter and trans-acting loss of function mutations in mgrB, which upregulate folA transcription in a PhoQP-dependent manner [44–46, 48]. Earlier work from our group identified Lon, a conserved quality control protease in bacteria, as a post-translational regulator of trimethoprim resistance. Lon degraded mutant DHFRs and reduced their half-lives in vivo, directly affecting genotype-phenotype relationships of resistance-conferring mutations [49, 50]. In this study, we first compare early mutational events in wild type and Lon-deficient E. coli challenged with trimethoprim. We show that a large genomic duplication encompassing folA confers trimethoprim resistance in both genetic backgrounds, though its frequency of occurrence is elevated in lon-knockout E. coli. We then use this system to ask what factors determine the fate of duplicated folA gene copies during evolution at different trimethoprim pressures.

Results

A large IS-flanked duplication encompassing folA mediates rapid adaptation to trimethoprim in E. coli

We had previously evolved 5 replicate populations of wild type E. coli K-12 MG1655 in sub-MIC trimethoprim (300 ng/mL) for 25 generations. We then isolated and sequenced the genome of a dominant trimethoprim-resistant clone from each population (designated as WTMPR1-5). These sequencing results have been reported earlier, in Patel and Matange (2021) [46]. Briefly, all isolates harboured loss-of-function mutations in the mgrB gene (Figure 1 A) that resulted in PhoQP-dependent up-regulation of DHFR expression [46]. Additionally, isolates WTMPR1, 3 and 5 had mutations at the active site of DHFR, while WTMPR4 harboured a large genomic duplication of ∼0.6 Mb encompassing the folA gene (Figure 1 A, B). We performed similar evolution experiments starting with an isogenic E. coli Δlon strain to ask if early adaptation to trimethoprim was altered by Lon-deficiency. Among trimethoprim-resistant isolates derived from E. coli Δlon, designated as LTMPR1-5, four harboured mutations at the mgrB locus (Figure 1 A). However, none of them had mutations in DHFR or its promoter. Instead, LTMPR2-4 had evolved GDA mutations similar to WTMPR4, while LTMPR1 had a duplication in an overlapping but shorter genomic stretch (Figure 1 A, B). In all trimethoprim-resistant isolates, duplicated genomic regions were flanked by IS-elements (Figure 1 B). Isolates with and without genomic duplications showed comparable trimethoprim IC50 values (Figure 1 A), indicating that GDA mutations were an alternate genetic path to evolving resistance and occurred more frequently in Lon-deficient E. coli.

Adaptation to antibiotics by large genomic duplications is favoured in lon-deficient E. coli.

A. Coverage depth plots and mutations at the folA and mgrB loci of 5 independently evolved trimethoprim-resistant isolates derived from wild type (WTMPR1-5) or Δlon E. coli (LTMPR1-5). Resistant clones were isolated after ∼25 generations of evolution at 300 ng/mL of trimethoprim. Coverage depth plots show the number of reads from Illumina short-read sequencing (y-axis) mapped to each genomic coordinate of the reference genome (x-axis). Coverage of 1x, 2x and 3x are marked by dotted lines for reference. Mutations in mgrB and folA, and trimethoprim IC50 values (mean ± S.D. from 3 independent measurements) are provided for each isolate below the appropriate coverage depth plot. B. Cartoon (not to scale) showing the duplicated genomic stretch and flanking IS-elements in trimethoprim-resistant isolates. Blue arrows represent IS1, while red arrows represent IS186. The positions of folA and acrAB genes, implicated in trimethoprim resistance in earlier studies, relative to the duplicated stretch are shown. Fold increase in the number of reads corresponding to the folA gene are indicated for each isolate. C. Summary of the prevalence of GDA mutations detected in wild type or E. coli Δlon after ∼25 generations of evolution in control media (no antibiotic) or in media supplemented with sub-MIC antibiotics. Each circle represents one of the 5 replicates in the evolution experiment. Light grey circles represent replicates in which no resistance was detected and hence were not sequenced. Dark grey circles indicate that the whole population or a resistant isolate was sequenced but no GDA mutation was detected. Red circles indicate that a GDA mutation was identified.

Seeking to explain the higher frequency of GDA mutations in the lon-knockout, we asked whether IS-transposases were sensitive to proteolysis by Lon. Indeed, heterologously expressed InsB and InsL had greater stability in the lon-knockout (Figure S1), suggesting that their accumulation may contribute to higher GDA frequency in E. coli Δlon. If true it would follow that a higher proportion of GDA mutants would be isolated from E. coli Δlon during adaptation to other environments as well. In agreement with this prediction, GDA mutations were detected after evolution of E. coli Δlon in 2 of the 5 additional growth conditions tested, namely antibiotic-free and spectinomycin-supplemented media (Figure 1 C, S2). In all cases duplications were flanked by IS elements (Figure S2). Further, the regions of the genome that were duplicated were similar among isolates derived from the same environment, but different across environments (Figure 1 A, S2). On the other hand, wild type E. coli did not show GDA mutations in any of the other tested conditions.

The duplicated genomic region in trimethoprim-resistant isolates included genes previously associated with resistance to trimethoprim, namely, folA which codes for DHFR, and acrA and B, which code for components of the AcrAB-TolC efflux pump (Figure 1 B). In one of the isolates, i.e. LTMPR1, the amplified stretch encompassed folA but not acrAB without affecting trimethoprim IC50 values (Figure 1 A, B). We therefore reasoned that duplication of folA was the main effector of adaptation to trimethoprim. Immunoblotting showed that DHFR protein was indeed overproduced in isolates LTMPR1 and WTMPR4 over and above the increase mediated by inactivation of mgrB alone (Figure 2 A).

Overexpression of DHFR in trimethoprim-resistant isolates with genomic duplication.

A. DHFR protein expression in trimethoprim-resistant isolates WTMPR4 and LTMPR1, compared with respective ancestors and mgrB-knockout strains. DHFR protein was detected by immunoblotting using an anti-DHFR polyclonal antibody. FtsZ was used as loading control. A representative immunoblot from 3 biological replicates is shown. Quantitation was performed by calculating band intensities using image analysis. Fold change values of DHFR levels over controls (wild type or E. coli Δlon, set to 1) is shown as mean ± S.D. from triplicate measurements. B. Expression level of duplicated genes in LTMPR1 and WTMPR4, determined by RNA-sequencing, expressed as fold over E. coli ΔlonΔmgrB and E. coli ΔmgrB respectively. Each point represents fold change value for a single gene. Mean of the scatter is shown as a black line, and its value is provided along with standard deviation above the plot. C, D. Correlation between expression levels of genes in LTMPR1 (C) and WTMPR4 (D) with E. coli ΔlonΔmgrB and E. coli ΔmgrB respectively. Gene expression levels were estimated using RNA-sequencing and fold changes were calculated using wild type E. coli as reference for WTMPR4 and E. coli ΔmgrB, and E. coli Δlon as reference for LTMPR1 and E. coli ΔlonΔmgrB. Each point represents log2(fold change) value for a single gene. The resulting scatter was fit to a simple linear regression and the obtained R2 values are provided. E. Relative fitness (w) of indicated E. coli strains calculated using competition experiments with E. coli ΔlacZ as reference. For strains derived from E. coli Δlon, E. coli ΔlonΔlacZ was used as reference. Neutrality of ΔlacZ was established by competition with the unmarked ancestral strains, i.e. wild type (WT) and E. coli Δlon. No change in relative fitness compared to the reference strain (w = 1) is indicated by a dotted line. Mean ± S.D. from three independent measurements is plotted.

Though DHFR overproduction may contribute to the selection of genomic amplifications in trimethoprim-resistant bacteria, ∼173 (LTMPR1) and ∼531 (WTMPR4) other genes were also duplicated in these isolates. To understand the global transcriptional burden imposed by this amplification, we analysed the gene expression profiles of WTMPR4 and LTMPR1 using RNA-sequencing (Supplementary File 1). Since both isolates had loss-of-function mutations in mgrB, which cause large changes in global gene expression [51], we also analysed the transcriptomes of E. coli ΔmgrB and E. coli ΔlonΔmgrB strains for comparison (Supplementary File 1). The level of folA mRNA was higher in LTMPR1 (2.3-fold) and WTMPR4 (5.7-fold) over respective controls (Supplementary File 1), in agreement with immunoblotting data. Other duplicated genes were mildly overexpressed by ∼1.7-fold on an average in LTMPR1 (over E. coli ΔlonΔmgrB) and ∼1.3-fold in WTMPR4 (over E. coli ΔmgrB) (Figure 2 B, Supplementary File 1). However, there was large variation in expression level among duplicated genes in these isolates (Figure 2 B), indicating that increase in gene dosage did not necessarily result in enhanced expression levels. Instead, global gene expression profiles of WTMPR4 and LTMPR1 were correlated with E. coli ΔmgrB and E. coli ΔlonΔmgrB respectively (Figure 2 C, D) and most of the top 50 up/down-regulated genes in WTMPR4 and LTMPR1 could be attributed to loss of mgrB alone (Figure S3). In line with this result, strains with and without genomic amplifications had comparable fitness costs (∼20 %) in drug-free media, which could also be attributed to inactivation of mgrB (Figure 2 E). Thus, we concluded that large GDA mutations encompassing folA led to DHFR overexpression and represented an alternate mechanism for trimethoprim-resistance in E. coli with undetectable additional cost.

Maintenance of folA duplication is linked to antibiotic-generated demand for DHFR expression

We next investigated the evolutionary fate of duplicate folA gene copies by evolving the LTMPR1 isolate, which harboured the shortest duplication, in the absence of antibiotic. We also evolved this isolate at sustained drug pressure, i.e. trimethoprim at the same concentration at which it had originally emerged (300 ng/mL). Using three replicate populations in each condition, we asked how DHFR expression and trimethoprim IC50 values changed over ∼252 generations of evolution and used genome sequencing to identify the underlying mutational basis. As controls, we performed similar evolution experiments with WTMPR4 (which harboured a GDA and a functional Lon protease) and WTMPR5 (which harboured a folA point mutation and a functional Lon protease).

Evolution in trimethoprim-free media

In drug-free conditions, trimethoprim IC50 of LTMPR1 rapidly declined and a concomitant loss of colony formation on trimethoprim-supplemented plates occurred (Figure 3 A, B). Evolved WTMPR4, but not WTMPR5, showed similar loss of colony formation on trimethoprim indicating that resistance mediated by GDA mutations was far less stable than point mutations regardless of the presence of Lon (Figure 3 B). For LTMPR1, these trends were explained by a decline in DHFR expression levels over evolution and reversion to a single folA copy (Figure 3 C, D, S4). Mutation in mgrB was, however, consistently maintained till the end of the experiment (Supplementary File 2). In line with these data, trimethoprim IC50s of evolving populations stabilized at a value similar to E. coli ΔlonΔmgrB (Figure 3 A). Interestingly, a second GDA mutation (“GDA-2”, ∼2.2 Mb to 3.3 Mb) emerged in these populations (Figure 3 D, S4). The GDA-2 mutation was apparently lost and regained multiple times during the experiment (Figure 3 D, S4), suggesting a dynamic maintenance in the population rather than fixation. Further, the region of the genome amplified in GDA-2 was similar to what we had observed when E. coli Δlon was evolved in drug-free media (Figure S2, S4), indicating that GDA-2 was likely a fitness enhancing mutation and unrelated to trimethoprim adaptation.

Evolutionary fate of folA-duplication in the absence of drug pressure.

A. Trimethoprim IC50 values of 3 evolving lineages derived from LTMPR1 (A, B, C) over 252 generations of evolution in antibiotic-free medium. Mean ± S.D. from three measurements is plotted at each time point. IC50 values of E. coli Δlon and ΔlonΔmgrB are shown as dotted lines for reference. B. Colony formation of LTMPR1, WTMPR4 and WTMPR5 evolved in antibiotic-free medium at different generations. Fraction of the population capable of forming colonies at 1, 5 and 10 μg/mL trimethoprim was calculated across 3 replicate lines. Mean value from the three lines is plotted. The results of similar experiments performed on ancestors (0 generations) are also provided for reference. C. DHFR expression during evolution in the absence of trimethoprim in LTMPR1 lineages A, B and C, measured by immunoblotting using anti-DHFR polyclonal antibody. FtsZ was used as a loading control. Quantitation was performed by calculating band intensities using image analysis. DHFR expression at each time point was normalized to the ancestor (i.e. 0 generations, set to 1). Mean of three independent measurements is shown below each lane. D. Copy number of the ancestral GDA encompassing folA (GDA(folA)) and GDA-2 at different time points of evolution in the absence of trimethoprim. For GDA(folA), copy number was determined by dividing number of reads from an Illumina sequencing experiment corresponding to folA by the average number of reads mapping to the rest of the genome. E. Point mutations in folA, rpoS and mgrB, and the “GDA-2” genomic duplication in 6 randomly picked colonies from each of the LTMPR1 lines at 252 generations of evolution are shown schematically using the appropriate symbols. Asterix (*/**) marks the genotypes that were carried forward for further analyses. F, G. Point mutations in folA, mgrB and rpoS in isolates derived from 252 generations of evolution of WTMPR4 (F) and WTMPR5 (G) in no antibiotic. From each of the evolving lines 2 random isolates were picked for genome sequencing. Various point mutants at the 3 gene loci of interest are represented by appropriate symbols as shown.

To further investigate the emergence of GDA-2 during evolution of LTMPR1 in drug-free media we sequenced 6 randomly picked colonies from each of the lineages at the final time point (Figure 3 E). All the sequenced isolates had a single copy of folA and a majority of isolates showed the presence of GDA-2, corroborating population sequencing results (Figure 3 E, Supplementary File 2). Interestingly, some of the isolates that did not harbour GDA-2 had evolved mutations in the rpoS gene (Figure 3 E, Supplementary File 2). Loss-of-function mutations in RpoS are fixed at late time points during adaptation to trimethoprim and enhance the fitness of E. coli without conferring resistance [46, 51]. We also sequenced a few randomly picked isolates from WTMPR4 and WTMPR5 evolved lines (2 from each of the 3 replicate lineages in each case) to check for the presence of GDA-2 and rpoS mutations. As expected, all WTMPR4 isolates had reverted to a single copy of folA, while WTMPR5 isolates retained their ancestrally inherited folA mutation. None of them had evolved the GDA-2 mutation, while a few of the isolates from WTMPR4 and WTMPR5 lines had acquired rpoS mutations (Figure 3 F, G).

Consistent with earlier studies, and the role of mutations in rpoS in general fitness enhancement in E. coli, we found that an LTMPR1-derived isolate with mutant rpoS recovered fitness in drug-free media (Figure 4 A). Interestingly, an isolate with the GDA-2 mutation had also recovered fitness (Figure 4 A). Coupled with the fact that the GDA-2 mutation occurred only in a lon-deficient background, and GDA-2 and rpoS mutations were mutually exclusive we wondered whether the fitness enhancing effects of GDA-2 were also mediated by reduced RpoS activity. Lon is known to reduce the levels of RpoS during the exponential growth phase of E. coli [52]. Therefore, in a lon-knockout we reasoned that at least some targets of RpoS would be overexpressed. To test this prediction, we analysed the transcriptome of E. coli Δlon and specifically looked at the expression levels of 168 known targets of RpoS (Supplementary File 3). Of these, 41 genes were overexpressed in the lon-knockout by at least 1.5-fold over wild type, including genes induced in the stationary phase and by various stresses such as poxB, aidB and cbpA (Figure 4 B, C). Levels of rpoS mRNA were however unaffected in this strain. Next, we compared the expression levels of these 41 RpoS-targets in LTMPR1 and E. coli Δlon ΔmgrB with E. coli Δlon as the reference and found that most had similar expression levels in these three strains (Figure 4 C). Indeed a few of these genes showed higher expression levels in LTMPR1 and E. coli Δlon ΔmgrB, which was not unexpected since loss of mgrB is known stabilize RpoS [53]. Finally, we checked their expression in isolates derived from LTMPR1 which harboured either GDA-2 or rpoS mutations. Indeed, in both isolates RpoS target genes were down-regulated indicating a re-setting of RpoS activity (Figure 4 C). Further, in the isolate harbouring GDA-2, rpoS mRNA levels were themselves lowered to 0.49x of E. coli Δlon (Supplementary File 3). Finally, deletion of rpoS from E. coli Δlon enhanced the fitness of the strain (Figure 4 D). Thus, evolution of LTMPR1 in antibiotic-free media was characterized by rapid reversal of GDA-mediated resistance and acquisition of mutations that enhanced fitness by reducing the levels of RpoS-transcribed genes (Figure 4 D).

Fitness enhancement of LTMPR1 in drug-free media results from lower RpoS activity.

A. Relative fitness (w) of isolates with the indicated genotypes derived from LTMPR1 evolution in antibiotic-free media (see also Figure 3) calculated using a competitive growth assay using E. coli ΔlonΔlacZ as the reference strain. Mean ± S.D. from three independent measurements is plotted. No change in relative fitness compared to the reference (w = 1) is shown as a dotted line for reference. B. Expression level of 168 known targets of RpoS in E. coli Δlon compared to wild type expressed as log2(fold change) values determined by RNA sequencing. Dotted lines at 1.5-fold higher and lower than wild type represent cut-offs used to identify overexpressed and under-expressed genes. The pie chart shows the fraction of RpoS targets that were overexpressed in E. coli Δlon by at least 1.5-fold in green. C. Expression levels of 41 RpoS targets in E. coli Δlon compared to wild type are shown as a heat map in the first vertical. The expression levels of these genes in LTMPR1, E. coli ΔlonΔmgrB or LTMPR1-derived isolates compared to E. coli Δlon are provided in the subsequent verticals. D. Relative fitness of E. coli ΔlonΔrpoS in antibiotic-free media calculated using a competitive growth assay using E. coli ΔlonΔlacZ as the reference strain. Neutrality of the ΔlacZ genetic marker was verified by competition between E. coli Δlon and E. coli ΔlonΔlacZ. Mean ± S.D. from three independent measurements is plotted. No change in relative fitness compared to the reference (w = 1) is shown as a dotted line for reference. E. Model for the effects of Lon deficiency on RpoS levels and bacterial fitness. The roles of mutations within RpoS or the GDA-2 mutations in compensating for the defects of RpoS overproduction are shown.

Evolution at constant trimethoprim pressure

In the presence of trimethoprim (300 ng/mL), antibiotic resistance level and DHFR expression remained high over the course of evolution of LTMPR1 (Figure 5 A, B, C). WTMPR4 and WTMPR5 too remained resistant, and in the case of WTMPR4 an increased colony forming efficiency was observed at a late time point (Figure 5 B). These data suggested that under drug pressure high folA copy number was likely to be maintained. Indeed, population sequencing of LTMPR1 lineages showed that the ancestral GDA mutation was maintained until 196 generations of evolution in all three replicates (Figure 5 D, S5). By 252 generations, however, a majority of bacteria in Lines A and C had reverted to a single folA copy (Figure 5 D, S5) and had acquired a point mutation either in the folA gene (Line C) or its promoter (Lines A) (Figure 5 D, Supplementary File 4). Line B, on the other hand, retained high folA copy number, but with a reduction in the length of the genomic duplication (Figure 5 D, S5). Line B had also acquired a mutation in the folA promoter, however its frequency in the population was low (Figure 5 D, Supplementary File 4).

Evolutionary fate of folA-encompassing GDA in LTMPR1 at sustained trimethoprim pressure.

A. Trimethoprim IC50 values of 3 evolving lineages starting from LTMPR1 (A, B, C) over 252 generations of evolution at 300 ng/mL trimethoprim. Mean ± S.D. from three measurements is plotted at each time point. IC50 values of E. coli Δlon and ΔlonΔmgrB are shown as dotted lines for reference. B. Colony formation of LTMPR1, WTMPR4 and WTMPR5 evolved in 300 ng/mL trimethoprim at indicated time points. Fraction of the population capable of forming colonies at 1, 5 and 10 μg/mL trimethoprim was calculated across 3 replicate lines. Mean value from the three lines is plotted. The results of similar experiments performed on ancestors (0 generations) are also provided for reference. C. DHFR expression in LTMPR1 lineages A, B and C evolved in 300 ng/mL trimethoprim, measured by immunoblotting using anti-DHFR polyclonal antibody. FtsZ was used as a loading control. Quantitation was performed by calculating band intensities using image analysis. DHFR expression at each time point was normalized to the ancestor (i.e. 0 generations, set to 1). Mean of three independent measurements is shown below each lane. D. Copy number of folA (GDA(folA)) at different time points of evolution of LTMPR1 in trimethoprim is plotted on the left y-axis. Copy number was determined by dividing number of reads from an Illumina sequencing experiment corresponding to folA by the average number of reads mapping to the rest of the genome. Frequency of various folA alleles in the evolving populations at each of the time points is also plotted on the right y-axis. E. Point mutations in folA, rpoS and mgrB in 6 randomly picked trimethoprim-resistant colonies from each of the LTMPR1 lines at 252 generations of evolution are shown diagrammatically. Genotypes that were carried forward for further analysis are marked (* or #). F, G. Point mutations in folA, mgrB and rpoS in isolates derived from 252 generations of evolution of WTMPR4 (F) and WTMPR5 (G) in 300 ng/mL of trimethoprim. From each of the evolving lineages 2 random isolates were picked for genome sequencing. Various point mutants at the 3 gene loci are represented by appropriate symbols as shown.

In agreement with population sequencing, none of the resistant isolates derived from LTMPR1 Line A at the final time point showed the ancestral GDA. Instead, they all harboured a C-35T mutation in the folA promoter that is known to up-regulate DHFR expression [51] (Figure 5 E, Supplementary File 4). Among Line B isolates, many had retained the ancestral duplication, without acquiring any additional mutations in folA. One of the isolates from this Line that had reverted to a single folA copy had gained the A-64T folA promoter mutation (Figure 5 E, Supplementary File 4). In Line C too we found that isolates that had lost the ancestral GDA had evolved a point mutation in the folA promoter (+TG-30) or a Trp30Arg mutation in DHFR (Figure 5 E, Supplementary File 4). We found similar results in sequenced isolates from WTMPR4 lineages, i.e. loss of GDA and gain of a point mutation either in folA or its promoter (Figure 5 F). WTMPR5-derived isolates did not accumulate additional mutations in folA but did acquire mutations in rpoS (Figure 5 G), showing that at this drug pressure additional mutations in folA were not required. Interestingly, two isolates from LTMPR1 Lines B and C had evolved new GDA mutations encompassing folA. In one of the Line B isolates (Isolate 2) the ancestral GDA was replaced by 5x amplification of a shorter genomic region (∼12 kb) encompassing folA (Figure S6). Both ends of this newly amplified region were flanked by IS1A (Figure S6). On the other hand, one of the Line C isolates (Isolate 5) had an expanded duplication beyond what was originally found in LTMPR1 (Figure S6). Also, both copies of folA in this isolate coded for DHFR-Trp30Arg (Figure 5 E, S6), suggesting that the original GDA was lost and an extended GDA mutation was gained after fixation of the mutant folA allele. Thus, even under drug pressure, in most bacteria GDA mutations were reversed. Notably, reversion to a single folA copy was slower than in drug-free media as it first required the fixation of point mutations at the folA locus. However, in a smaller sub-population, more than one folA gene copy was maintained. This maintenance was dynamic and likely involved repeated loss and gain of GDA mutations.

To further investigate the replacement of GDA by point mutants, we checked to see if under drug pressure point mutants could invade a population harbouring GDA mutations. We allowed LTMPR1-derived isolates with point mutations in folA (marked with ΔlacZ::Kan) to compete with a GDA-harbouring isolate from the same time point in evolution (to nullify the effects of other media adaptations) and asked how rapidly the former could invade a population of the latter. Since for this assay we used point mutant strains genetically marked with ΔlacZ::Kan we verified the neutrality of the ΔlacZ::Kan marker by checking if the marked isolate could invade its unmarked parent. Indeed, in the presence of trimethoprim (300 ng/mL) point mutants were enriched over the GDA mutant in co-culture by about 100-fold (Figure 6 A) recreating the results of our evolution experiment. A possible explanation for this finding may be higher resistance level conferred by folA point mutations than GDA. However, most isolates with folA point mutations or harbouring GDAs had similar IC50 values as the ancestral LTMPR1 (Figure 6 B). These isolates also had comparable doubling times in the presence of 300 ng/mL trimethoprim (Figure 6 B). Thus, though there were no large observable differences in growth rate or drug IC50 values between strains, point mutants could indeed invade and outcompete GDA mutations, reflecting either small differences in fitness between them or their relative stabilities in culture.

Invasion of GDA-dominant populations by point mutations under trimethoprim pressure.

A. Upper panel. Competition between isolates with point mutations in folA and an isolate that harboured a folA duplication derived from the same time point of LTMPR1 evolution (upper panel). The genotypes of the isolates used are shown diagrammatically (see also Figure 5, Supplementary File 4). The initial mixing ratio was 1000:1 in favour of the GDA mutant. Point mutants were marked genetically with ΔlacZ::Cat. Neutrality of the ΔlacZ::Cat marker was verified by competing marked and unmarked point mutants (lower panel). Percentage of the population constituted by the marked point mutants over ∼105 generations of serial transfer in media supplemented with trimethoprim (300 ng/mL) is plotted. Traces for individual mutants are shown in grey. Average of the 4 point mutants is shown in red. B. Trimethoprim IC50 values of the LTMPR1 ancestor or evolved isolates from 252 generations in trimethoprim with indicated genotypes (see also Figure 5, Supplementary File 4) are plotted as bars. Mean ± S.D. from three independent measurements are plotted. Doubling times relative to LTMPR1 (set to 1) in the presence of 300 ng/mL of trimethoprim are provided above the graph as mean ± S.D. from three independent measurements.

Maintenance of high folA gene copy number under drug pressure is driven by unstable mutants of DHFR

Though sustained drug pressure did extend the lifetime of folA gene duplication, acquisition of trimethoprim-resistant point mutations facilitated reversion to a single gene copy. We therefore asked whether folA duplication could be maintained under conditions of increasing antibiotic pressure, i.e., trimethoprim concentrations starting at 1 μg/mL and ending at 10 μg/mL over 252 generations of evolution (Figure 7 A). Under these conditions, trimethoprim IC50 values increased sharply over the course of evolution eventually reaching ∼40 to ∼160-fold higher than ancestral LTMPR1 (Figure 7 A). Colony formation at high trimethoprim concentrations also increased at later generations of evolution (Figure 7 B), reflecting the high drug pressure. Similarly evolved WTMPR4 and WTMPR5 also showed enhanced colony formation at high trimethoprim concentrations (Figure 7 B).

Evolutionary fate of folA-encompassing GDA in LTMPR1 at increasing trimethoprim pressure.

A. Trimethoprim IC50 values of 3 evolving lineages starting from LTMPR1 (A, B, C) over 252 generations of evolution are plotted on the left Y-axis. Mean ± S.D. from three measurements is plotted at each time point. Trimethoprim concentrations used during evolution are plotted in the right Y-axis. B. Colony formation of LTMPR1, WTMPR4 and WTMPR5 evolved in increasing trimethoprim at indicated time points. Fraction of the population capable of forming colonies at 1, 5 and 10 μg/mL trimethoprim was calculated across 3 replicate lines. Mean value from the three lines is plotted. The results of similar experiments performed on ancestors (0 generations) are also provided for reference. C. DHFR expression in LTMPR1 lineages A, B and C evolved in increasing trimethoprim, measured by immunoblotting using anti-DHFR polyclonal antibody. FtsZ was used as a loading control. Quantitation was performed by calculating band intensities using image analysis. DHFR expression at each time point was normalized to the ancestor (i.e. 0 generations, set to 1). Mean of three independent measurements is shown below each lane. D. Copy number of folA (GDA(folA)) at different time points of LTMPR1 evolution in increasing trimethoprim is plotted on the left y-axis. Copy number was determined by dividing number of reads from an Illumina sequencing experiment corresponding to folA by the average number of reads mapping to the rest of the genome. Frequency of various folA alleles in the evolving LTMPR1 populations at each of the time points is shown on the right y-axis. E. Point mutations in folA, rpoS and mgrB in 6 randomly-picked trimethoprim-resistant colonies from each of the LTMPR1 lines at 252 generations of evolution are shown schematically. F, G. Point mutations in folA, mgrB and rpoS in isolates derived from 252 generations of evolution of WTMPR4 (F) and WTMPR5 (G) in increasing trimethoprim. From each of the evolving lineages 2 random isolates were picked for genome sequencing. Various point mutants at the 3 gene loci are represented by appropriate symbols as shown.

Curiously, we observed that the 3 replicate lineages of LTMPR1 showed different trends in DHFR expression. Line A maintained a high level of DHFR protein for the duration of the experiment. On the other hand, Lines B and C showed high DHFR expression until 196 generations, after which it sharply declined (Figure 7 C). Genome sequencing revealed that all 3 lines convergently evolved the Trp30Arg mutation in DHFR early during evolution and had fixed this mutation by 196 generations. All 3 lines also retained folA duplication until this time point (Figure 7 D, S7). Subsequently, Line A evolved an additional C-35T folA promoter mutation and had partially lost the ancestral GDA (Figure 7 D, S7, Supplementary File 5). In Lines B and C, a second missense mutation in DHFR, i.e. Tyr151Asp, appeared between 196 and 252 generations (Figure 7 D, Supplementary File 5). Line B also showed an Ile94Leu mutation in DHFR at low frequency at generation 252 (Figure 7 D, Supplementary File 5). Unexpectedly, despite acquiring these additional mutational hits and lower DHFR expression level, bacteria in LTMPR1 Lines B and C retained two copies of folA in contrast to Line A (Figure 7 D, S7, Supplementary File 5).

Individual resistant clones from each of the lineages provided greater insight into the genetic trajectories of evolution. In LTMPR1 Line A, 2 of the 6 sequenced isolates had lost gene duplication and acquired the Trp30Arg coding and C-35T promoter mutations in folA (Figure 7 E, Supplementary File 5). Interestingly, the other 4 sequenced isolates from Line A retained the ancestral GDA and harboured the same combination of folA point mutations, but only in 1 of the 2 gene copies (Figure 7 E, Supplementary File 5). Ten of the twelve isolates from LTMPR1 Lines B and C, on the other hand, retained folA duplication. In all ten isolates, both copies of folA had acquired Trp30Arg and Tyr151Asp mutations (Figure 7 E, Supplementary File 5). Unlike LTMPR1, all sequenced isolates of WTMPR4 at 252 generations of evolution had lost their GDA mutation and gained 2 folA mutations (Figure 7 F). WTMPR5 too had gained an additional folA mutation (Figure 7 G). Notably, none of the isolates from WTMPR4 and WTMPR5 lineages had the Trp30Arg and Tyr151Asp combination of mutations that occurred repeatedly in LTMPR1 lineages.

We next sought to understand why some LTMPR1 bacteria had retained high folA copy number despite acquiring multiple point mutations in the drug target. To address this question, we first asked what the advantage of folA duplication was in the two different mutational trajectories that were observed for LTMPR1. Isolates from Line A had similar IC50 values for trimethoprim regardless of whether they had one or two copies of folA (Figure 8 A). On the contrary, isolates from Line B that harboured only one copy of mutant folA (coding for DHFR-Trp30Arg-Tyr151Asp) displayed ∼3.5-fold lower IC50 than those that harboured two copies (Figure 8 A). Immunoblotting showed that isolates from Line A had similar expression level of DHFR (Figure 8 B). However, DHFR protein level was significantly compromised by the Trp30Arg-Tyr151Asp mutation combination and double the number of folA gene copies restored this defect (Figure 8 B). Thus, higher gene dosage of DHFR was dispensable once a promoter up-regulatory mutation was fixed. On the other hand, bacteria with alleles of DHFR harbouring multiple missense mutations continued to benefit from overexpression due to folA duplication.

Proteostatic pressure facilitates maintenance of folA duplication

A. Trimethoprim-IC50 values of resistant isolates with the indicated genotypes (see also Figure 7) derived from LTMPR1 evolution in increasing antibiotic pressure. Mean ± S.D. from three independent measurements is plotted. B. DHFR expression level in LTMPR1-derived trimethoprim resistant isolates with indicated genotypes assessed by immunoblotting using anti-DHFR polyclonal antibody. FtsZ was used as a loading control. Quantitation was performed by calculating band intensities using image analysis. DHFR expression was normalized to the ancestor (LTMPR1). Mean of three independent measurements is provided. C. Cartoon representation of the structure of E. coli DHFR (PDB: 7DFR) bound to folate. Residues Trp30, Tyr151, Phe153 and Ile155 which form hydrophobic interactions and are required for proteolytic stability are shown as sticks and coloured by element (C: green, O:red, N:blue). Distances of less than or equal to 4 Å, which indicates possible interactions, are shown as dotted yellow traces. Folate bound in the active site of DHFR is shown as sticks. D. Trimethoprim IC50 values of E. coli wild type or E. coli Δlon heterologously expressing DHFR (wt) or its mutants. Mean ± S.D. from three independent measurements is plotted. E. Expression level of plasmid-borne DHFR (wt) or its mutants in wild type or Δlon E. coli assessed by immunoblotting using anti-DHFR polyclonal antibody. FtsZ was used as a loading control. Quantitation was performed by calculating band intensities using image analysis. Expression of mutants was normalized to wild type DHFR (set to 1). Mean of three independent measurements is shown.

Mutations at position Trp30 in DHFR are accompanied by lower in vivo proteolytic stability due to perturbation of the hydrophobic core of the protein [54]. This, in turn, affects the expression level of DHFR [49, 54]. Interestingly, residue Tyr151 lies in the same hydrophobic pocket as Trp30 (Figure 8 C). To test the effect of combination of mutations at these sites on trimethoprim resistance and DHFR expression, we first heterologously expressed wild type DHFR, Trp30Arg or Tyr151Asp single mutants and the Trp30Arg-Tyr151Asp double mutant in wild type E. coli. As expected, DHFR Trp30Arg expression increased the IC50 of trimethoprim. Surprisingly, neither DHFR Tyr151Asp, nor the Trp30Arg-Tyr151Asp double mutant conferred resistance to the antibiotic. Indeed, both mutants conferred lower IC50 values than wild type DHFR. This effect was due to loss of hydrophobicity at the Tyr151 position, as DHFR-Tyr151Leu and DHFR-Tyr151Phe conferred similar IC50 values as wild type DHFR (Figure 8 D). Since the Trp30Arg-Tyr151Asp mutation combination had evolved in a lon-deficient background, we next expressed these mutants in E. coli Δlon. Here too, DHFR Trp30Arg conferred trimethoprim resistance, while the Tyr151Asp mutant did not. Strikingly, the combination of Trp30Arg and Tyr151Asp conferred very high-level trimethoprim resistance (Figure 8 D) demonstrating synergistic epistasis that was contingent on the absence of Lon protease. The levels of expression of DHFR mutants in wild type and lon-deficient bacteria mirrored trimethoprim resistance levels indicating that clearance by Lon altered their genotype-phenotype relationship (Figure 8 E). Based on these data, we concluded that the action of proteostatic machinery generated additional demand for expression by clearing mutant-DHFR proteins. This additional demand drove the maintenance of folA duplications in bacteria evolving at very high drug pressures.

Discussion

Ohno’s hypothesis, proposed in 1970, provided a conceptual framework to think about the evolution and eventual fate of duplicated genes [55]. Several theoretical and experimental studies have built upon the original idea to delineate the evolutionary outcomes of gene duplication [18, 56, 57]. More recently, experimental evolution of microbes has been used to test some of the suggestions made by Ohno and others. This approach has the inherent advantage of greater control over the selection pressure applied and available genetic tools to interrogate molecular mechanisms. In most studies of this kind, an artificial system was constructed which was then evolved under defined selection pressures [33, 58–61]. Our study began with serendipitous isolation of spontaneous trimethoprim-resistant mutants of E. coli that overproduced DHFR because of a folA gene duplication. We exploited this system to understand the relationship between selection pressure and gene duplication in the context of drug-resistance evolution. The selection pressure for higher gene expression, i.e. “expression demand”, can be understood as the amount of a gene product that is required to maintain fitness in a particular environment. Antibiotics directly increase expression demand by inhibiting the activity of their targets. Our study shows that increasing gene dosage is a rapid, but transient evolutionary response to ensure that expression demand is met. On the other hand, point mutations are relatively stable routes to mitigating expression demand. Over short time scales, changes in gene dosage can counter the effects of antibiotic. However, their inherent instability results in almost inevitable replacement with phenotypically equivalent point mutations. Thus, continued antibiotic exposure delayed but could not prevent reversion to low gene copy number in most of our experiments. Further, we discover a novel association between proteostasis and gene dosage evolution. In E. coli evolving at high trimethoprim pressure, bacterial protein quality control exacerbated expression demand by clearing mutant DHFR-s. As a result, drug-resistant mutants of DHFR with low proteolytic stability drove the maintenance of folA gene duplication (Figure 9). Based on these findings we propose that a combination of factors, which include environmental and physiological, alter the relationship between gene copy number and fitness, and drive gene dosage evolution.

Model of expression demand induced selection of folA gene duplication, followed by replacement by phenotypically equivalent point mutants.

Expression demand for DHFR is generated by trimethoprim pressure, which results in the selection of folA gene duplications. Gene duplication is unstable and reverses to a single copy of wild type folA when drug-pressure is withdrawn. Under drug pressure, folA gene duplication is maintained until a point mutation that confers resistance arises in one of the copies of folA (promoter or coding region). Upon acquisition of a point mutation, reversal of gene duplication ensues unless additional expression demand is generated by the action of proteostatic machinery on unstable drug-resistant DHFR mutants.

A key finding from our study was the two-fold influence that proteostatic machinery had on gene copy number. Firstly, Lon protease-deficiency promoted GDA mutations, possibly by allowing accumulation of IS-tranposases in E. coli. This observation resonates with earlier work by Nicoloff and co-workers who found that large IS-element flanked genomic duplications in tetracycline-resistant E. coli were commonly preceded by spontaneous mutations that inactivated lon [62, 63]. Further, the role of Lon protease in regulating IS-element mobility by a similar mechanism as proposed by us has been previously demonstrated for IS1 and IS903, strengthening our argument [64, 65]. We note however that other mechanisms by which Lon may modulate the frequency of IS-mediated GDAs in E. coli cannot be ruled out. For instance, Lon protease homologs from several organisms directly bind DNA [66–70], though the role of this binding in DNA transactions is unclear. Similarly, Lon is a regulator of the recombination protein, RecA, by its action on LexA [71] and in Salmonella RecA stimulated the rate of gene duplications mediated by IS3 [23]. The second impact of protein quality control on GDA was the action of proteases on trimethoprim-resistant mutants of DHFR, which promoted gene duplication by generating additional demand for expression (Figure 9). Importantly, which of these two effects dominated was determined by a combination of environment, i.e. drug concentrations, and other mutations acquired at the folA locus during evolution. In particular, mutations in the folA promoter could alleviate expression demand and hence reduce selection for high gene copy number.

The interplay between GDA and point mutations during evolution agrees in many respects with a recent study by Tomanek and Guet [61]. They use an elegantly designed artificial system [33, 61] in which expression of the galK gene in E. coli was driven by a very low activity promoter. This strain was then allowed to evolve in different concentrations of galactose. Like the present study, they found that the activity of IS-elements biased evolutionary trajectories in favor of duplications. They also showed that co-occurrence of duplications and point mutations was selected only at high expression demand, in agreement with our findings. Finally, they argued that gene duplication events may constrain evolution by point mutations. Once again, our study corroborates these results at the short term in that the lon-knockout which showed a higher propensity for GDA mutations did not evolve point mutations in folA at early time points of evolution. However, given the relative low stability of large GDAs, these were inevitably lost over longer duration of evolution and replaced with point mutations. Thus, it is likely that the different mutational profiles of IS-biased populations may be relevant only at short durations of evolution. It is important to note that though we have focused on the effects of mutations at the folA locus, which are directly under trimethoprim pressure, many of the isolates from our evolution experiments also harboured additional mutations at other loci. For instance, mutations in rpoS were seen throughout our evolution lines and at least in drug-free media were significant contributors to bacterial fitness. It is not yet clear what the role of other mutations are likely to be under antibiotic selection, though it may be envisaged that these additional mutations play the role of media adaptations. We did not take their effects into in this study account since we expected them to be minor contributors to fitness especially during evolution under trimethoprim selection. However, their role in modulating the stability and selection of GDA mutations in drug-free/low drug environments may be more significant and remains to be investigated in the future.

In summary, the present study uncovers the complex relationship between gene copy number, point mutations and environment, and identifies proteostasis as a regulator of gene dosage evolution in bacteria. These findings open the door to alternative explanations for why gene duplications may be maintained over long periods of evolution and provides insight into how gene expression demand is mitigated by multiple mechanisms in bacteria.

Materials and methods

Strains, plasmids and culture conditions

Escherichia coli K-12 MG1655 was used as wild type for all experiments. E. coli DH10B was used for plasmid construction and manipulation. E. coli strains were cultured in Luria-Bertani (LB) broth with shaking at 180-200 rpm or on LB agar plates at 37 °C. Ampicillin was used at a concentration of 100 μg/mL for selection of plasmid-transformed strains. Trimethoprim was added to growth media at required concentrations as needed. The strains and plasmids used in this study are listed in Table 1.

Laboratory evolution of antibiotic resistance

Short-term evolution of antibiotic resistance

For short term laboratory evolution of antibiotic resistance, wild type and E. coli Δlon populations were cultured in 96-well plates in LB medium incorporating sub-inhibitory concentrations of different antibiotics. Following concentrations of antibiotics were used: trimethoprim (300 ng/mL; MIC/3); spectinomycin (5 μg/mL; MIC/10); nalidixic acid (4.25 μg/mL; MIC/3); erythromycin (10 μg/mL; MIC/10); rifampicin (4.25 μg/mL; MIC/3). Control populations were also set up in antibiotic-free LB. Five replicates were included for each strain in each antibiotic. Total volume of culture in each well was 150 μL. 10 % of the population was passaged to fresh growth media every 24 hours, for 7 days, resulting in ∼25 generations of evolution. At the end of 7 days, the evolved populations were frozen as glycerol stocks.

To check for the evolution of resistance and isolation of resistant clones for further analyses, frozen populations were revived in drug-free LB broth and grown overnight (14-16 hours). Saturated cultures were serially diluted and spotted on LB agar plates supplemented with the appropriate antibiotic at MIC. Colonies were randomly picked, cultured overnight in drug-free LB and frozen in 50% glycerol at -80 °C until further use.

Long-term evolution at different trimethoprim pressures

For long-term evolution, trimethoprim-resistant isolates LTMPR1, WTMPR4 and WTMPR5 were first revived from frozen stocks in drug-free LB overnight. Triplicate evolving lineages were then established by sub-culturing 1% of the saturated culture of each ancestor into 2 mL of appropriate medium. For relaxed selection drug-free LB was used. For constant antibiotic pressure, trimethoprim was added at a concentration of 300 ng/mL. For increasing antibiotic pressure, trimethoprim was increased from 1 μg/mL to 10 μg/mL over course of the experiment as shown in Figure 4 A. Cultures were grown for 10-14 hours (∼7 generations) and then passaged (1%) into fresh media. This was continued for 36 passages, i.e. ∼252 generations. Aliquots of cultures were frozen periodically for further analyses.

For isolating individual clones, populations were revived from frozen stocks at 252 generations of evolution in drug-free LB, serially diluted and then spotted onto LB agar plates supplemented with trimethoprim (1, 5 or 10 µg/mL). Plates were incubated for 18-20 hours at 37 °C and resistant colonies were picked randomly for further analyses.

Measuring trimethoprim resistance

Trimethoprim IC50

Trimethoprim Inhibitory Concentration - 50 (IC50), i.e. the antibiotic concentration needed for 50 % growth inhibition, was calculated using a broth dilution assay. Frozen glycerol stocks of appropriate bacterial strains were revived in 1-3 mL LB for 6-8 hours. In a 96-well plate, 150 μL of LB was added to individual wells. Trimethoprim stocks were added, and the antibiotic was serially diluted. No antibiotic was added to a few wells, which served as controls. Next, 1.5 μL of the appropriate bacterial culture was added to each well. The peripheral wells of the plate were filled with water to prevent dehydration. The 96-well plate was then incubated at 37℃ for 15-18 hrs. Optical density at 600 nm was measured for each well using a plate reader (Ensight, Perkin Elmer). Obtained data were first normalised to growth in drug-free wells (set to 1) and then fitted to a sigmoidal log(inhibitor)-response curve using Graphpad Prism software. The equation for this model is as follows: where, Min: Minimum value of normalised OD and Max: maximum value of normalised OD Values of IC50 were estimated from the above curve fit and averaged across 3 independent measurements for every strain.

Colony forming efficiency

To calculate the colony forming efficiency of evolving populations at different trimethoprim concentrations, frozen glycerol stocks were first revived in 1 mL LB for 14-15 hours. Saturated cultures were serially diluted (10-fold). Diluted cultures were then spotted (10 mL) on LB agar plates containing 0, 1, 5 and 10 μg/mL of trimethoprim. Spots were allowed to dry completely and the plates were incubated at 37 °C for ∼18 hours. Colonies were counted from the least dense spot and multiplied by the appropriate dilution factor to determine colony forming units (CFUs)/mL.

Growth curves

Appropriate frozen stocks were revived overnight in drug-free LB (1-3 mL) and 1μl of the saturated culture was added to 200 μL of media supplemented with trimethoprim (300 ng/ml) in the wells of a sterile uncoated 96-well plate. Optical density at 600 nm was measured during growth at 37 °C at 15-minute intervals over a 24-hour period on a Perkin-Elmer Ensight Multimode Plate Reader. Doubling time (DT) (in mins) was estimated from log phase, using the following formula: Where OD1 and OD2 represent optical densities at the onset and end of log phase respectively, and T1 and T2 represent time elapsed (in mins), corresponding to OD1 and OD2.

Genome sequencing, identification of mutations and detection of GDA mutations

Genomic DNA was extracted from late log-phase cultures of appropriate strain/resistant isolates/evolving populations using phenol:chloroform:isoamyl alcohol extraction as previously described [72]. Extracted genomic DNA was then cleaned-up using HiGenoMB DNA isolation and purification kit (Himedia, India) as per manufacturer’s instructions and quantitated spectrophotometrically. The integrity was checked by running purified DNA on an agarose gel before further analyses.

For whole-genome resequencing, paired-end whole-genome next-generation sequencing (NGS) was performed on a MiSeq system (Illumina, USA) with read lengths of 150-250 bps. Library preparation and sequencing services were provided by Eurofins (Bangalore, India) and Genotypic (Bangalore, India). Processed reads were aligned to the reference genome E. coli K-12 MG1655 (U00096) using bowtie2. Variant calling and prediction of new junctions in the genome to identify large structural mutations was done using Breseq [73] using default settings in population or clonal modes as needed. All variants and new junctions that were already present in the ancestor were excluded. Variants unique to the test sample were classified as “mutations”.

For detection of GDA mutations, ‘coverage’ vs ‘coordinate in reference genome’ plot generated by the bam.cov command in Breseq was used [73]. Contiguous stretches greater than 10 kb with coverage depth of 1.5x or more, compared to the rest of the genome, were classified as GDA mutations. The presence of IS-elements or repeat sequences was manually checked using the coordinates of the GDA.

Calculating relative fitness of E. coli strains

Relative fitness of gene knockouts or evolved clones was estimated using competition assays. Saturated cultures of test strains and reference strains (E. coli ΔlacZ or E. coli ΔlacZΔlon) were mixed 1:1 to a total of 1% in 3 mL LB. Strains were allowed to compete for ∼7 generations (24 hours incubation). Mixed cultures were diluted and plated on LB agar supplemented with IPTG (1 mM) and X-gal (50 μg/mL) at the start and end of the competition experiment. Plates were incubated for 18-24 hrs, and blue and white colonies were counted to determine the CFUs/mL at initial and final time points.

Relative fitness (w) was calculated using the following formula: where Tf and Ti are the final and initial CFUs/mL of the test strain and Rf and Ri are the final and initial CFUs/mL of the reference strain, respectively.

Estimation of DHFR expression by immunoblotting

Immunoblotting was used to determine the expression level of DHFR in various strains/mutants of E. coli as described in Patel and Matange (2021) [46]. Briefly, equal lysate protein (5 μg) was subjected to SDS-PAGE on a 15% gel and electroblotted onto PVDF membranes. The membrane was blocked with 5% BSA solution and cut into 2 halves at the 25 kDa molecular weight band. The lower part of the membrane was probed with anti-DHFR polyclonal IgG (100 ng/mL) and the upper part was probed with anti-FtsZ serum (1:50,000). HRP-linked anti-rabbit IgG (1:10,000) was used as the secondary antibody. Hybridised antibodies were detected using chemiluminescence. Band intensities were quantitated in ImageJ and normalised to appropriate control (set to 1). Fold change in DHFR was calculated as follows:

Gene expression analysis using RNA-sequencing

Transcriptome analyses were performed using RNA sequencing. For RNA-seq, saturated overnight culture of appropriate strains/isolates was inoculated (1 %) into 3 mL LB in triplicate and incubated at 37 °C for 3 hours with shaking at 180 rpm. The cultures were centrifuged and bacterial pellets were resuspended in 1 mL of RNAlater (Invitrogen) for storage. RNA extraction, sequencing and preliminary data analyses were performed by Eurofins (Bangalore, India). Total RNA was extracted using RNAeasy spin columns and quantitated using Qubit and Bioanalyzer. Bacterial rRNA was depleted using Ribozero kit. Bacterial mRNA was then reverse transcribed, library prepared and quality control performed using Tapestation platform. Paired end sequencing was performed on Illumina platform with read lengths of 150 bp read length. Processed reads were aligned to the reference genome (U00096.3) using HISTAT2(version 2.1.0). Abundance estimation was done using featureCounts( version1.34.0). Differential gene expression (DGE) analysis was done by comparing the expression of individual genes in test and reference samples. The output of the DGE analyses were log2(fold change) values for each gene and P-values for statistical significance.

Cloning and site-directed mutagenesis

Cloning of InsB and InsL transposases

InsB was amplified from E. coli K-12 MG1655 genomic DNA by PCR using primers insB1_fwd_NcoI (5’-GGCGCCATGGATATGCCGGGCAACAGCCCG-3’) insB1_rev_EcoRI (5’-CCTTTGAATTCTTATTGATAGTGTTTTATG-3’). For amplification of insL, primers insL_EcoRI_fwd (5’-ATGCCGAATTCATGAATTACTCTCACGATAAC-3’) and insL_HindIII_rev (5’-ATGCCAAGCTTTTAGTTCTTCTTTTCGGATCC-3’) were used. Amplified insB gene was digested with NcoI and EcoRI restriction endonuclease and ligated with similarly digested pPROEx-HtC plasmid. Amplified insL was digested with EcoRI and HindIII restriction endonucleases and ligated with similarly digested pPROEx-HtA plasmid. Positive clones were screening using restriction digestion. Clones were transformed into E. coli wild type or Δlon for further analyses.

Site-directed mutagenesis for DHFR

Site-directed mutagenesis was performed to introduce mutations into the coding sequence of folA in pPRO-folA as described in [54]. Briefly, PCR was performed using pPRO-folA or pPRO-folA Trp30Arg as template and the appropriate forward and reverse mutagenic primers given below:

folA_Tyr151Asp_fwd_BglII:

5’-CTCTCACAGCGATTGCTTTGAGATCTTGGAGCGGCGGGGCC-3’

folA_Tyr151Asp_rev_BglII:

5’-CGCCGCTCCAAGATCTCAAAGCAATCGCTGTGAGAGTTCTGCG-3’

folA_Tyr151Leu_fwd_BglII:

5’-TCTCACAGCTTATGCTTTGAGATCTTGGAGCGGCGGGGCC-3’

folA_Tyr151Leu_rev_BglII:

5’-CGCCGCTCCAAGATCTCAAAGCATAAGCTGTGAGAGTTCTGCG-3’

folA_Tyr151Phe_fwd_BglII:

5’-CTCACAGCTTTTGCTTTGAGATCTTGGAGCGGCGGGGCC-3’

folA_Tyr151Phe_rev_BglII:

5’-GCCGCTCCAAGATCTCAAAGCAAAAGCTGTGAGAGTTCTGCG-3’

PCR products were digested with DpnI overnight, transformed into chemically competent E. coli DH10B and selected on LB agar plates containing ampicillin. Plasmids habouring the desired mutations were screened by restriction digestion and confirmed by sequencing. Sequenced clones were transformed into E. coli wild type or Δlon for further analyses.

In-cell stability of InsB and InsL

A chloramphenicol-chase assay was used to assess the in-cell stability of heterologously expressed InsB and InsL transposases as described in Matange (2020) [49] with a few modifications. Overnight cultures of E. coli wild type or Δlon harbouring plasmids pPRO-insB or pPRO-insL were diluted 1% in 5 mL LB supplemented with ampicillin and IPTG (0.1mM). After 4 hours of growth at 37 °C, 1 mL of culture was pelleted. The cell pellet was lysed in 100 μL of 1x Laemmli sample loading buffer by heating at 95°C for 10-20 mins. Chloramphenicol was added to the remaining culture at a final concentration of 25 μg/mL to stop protein synthesis and the culture was placed back in the incubator. After 15, 45, and 90 minutes 1 mL aliquots of the culture were pelleted and lysed. Lysates were subjected to SDS-PAGE followed by immunoblotting to detect the levels of InsB and InsL. Anti-Hexahis monoclonal antibody (Invitrogen) was used as the primary antibody for InsL. Anti-Hexahis polyclonal antibody (Invitrogen) was used as the primary antibody for InsB. Band intensities were calculated in ImageJ and normalised to control (set to 1) to calculate fraction of protein remaining after chloramphenicol treatment.

eLife Assessment

This important study explores the interplay between gene dosage and gene mutations in the evolution of antibiotic resistance. The authors provide solid evidence to connect proteostasis with gene duplication during experimental evolution in a model system. If the experiments are found to be rigorous and reproducible, then this paper will be of high interest to other researchers studying antibiotic resistance, proteostasis, and bacterial evolution.

Reviewer #1 (Public review):

Summary:

The study by Jena et al. addresses important questions on the fundamental mechanisms of genetic adaptation, specifically, does adaptation proceed via changes of copy number (gene duplication and amplification "GDA") or by point mutation. While this question has been worked on (for example by Tomanek and Guet) the authors add several important aspects relating to resistance against antibiotics and they clarify the ability of Lon protease to reduce duplication formation (previous work was more indirect).

A key finding Jena et al. present is that point mutations after significant competition displace GDA. A second one is that alternative GDA constantly arise and displace each other (see work on GDA-2 in Figure 3). Finally, the authors found epistasis between resistance alleles that was contingent on lon. Together this shows an intricate interplay of lon proteolysis for the evolution and maintenance of antibiotic resistance by gene duplication.

Strengths:

The study has several important strengths: (i) the work on GDA stability and competition of GDA with point mutations is a very promising area of research and the authors contribute new aspects to it, (ii) rigorous experimentation, (iii) very clearly written introduction and discussion sections. To me, the best part of the data is that deletion of lon stimulates GDA, which has not been shown with such clarity until now.

Weaknesses:

The minor weaknesses of the manuscript are a lack of clarity in parts of the results section (Point 1) and the methods (Point 2).

Reviewer #2 (Public review):

Summary:

In this strong study, the authors provide robust evidence for the role of proteostasis genes in the evolution of antimicrobial resistance, and moreover, for stabilizing the proteome in light of gene duplication events.

Strengths:

This strong study offers an important interaction between findings involving GDA, proteostasis, experimental evolution, protein evolution, and antimicrobial resistance. Overall, I found the study to be relatively well-grounded in each of these literatures, with experiments that spoke to potential concerns from each arena. For example, the literature on proteostasis and evolution is a growing one that includes organisms (even micro-organisms) of various sorts. One of my initial concerns involved whether the authors properly tested the mechanistic bases for the rule of Lon in promoting duplication events. The authors assuaged my concern with a set of assays (Figure 8).

More broadly, the study does a nice job of demonstrating the agility of molecular evolution, with responsible explanations for the findings: gene duplications are a quick-fix, but can be out-competed relative to their mutational counterparts. Without Lon protease to keep the proteome stable, the cell allows for less stable solutions to the problem of antibiotic resistance.

The study does what any bold and ambitious study should: it contains large claims and uses multiple sorts of evidence to test those claims.

Weaknesses:

While the general argument and conclusion are clear, this paper is written for a bacterial genetics audience that is familiar with the manner of bacterial experimental evolution. From the language to the visuals, the paper is written in a boutique fashion. The figures are even difficult for me - someone very familiar with proteostasis - to understand. I don't know if this is the fault of the authors or the modern culture of publishing (where figures are increasingly packed with information and hard to decipher), but I found the figures hard to follow with the captions. But let me also consider that the problem might be mine, and so I do not want to unfairly criticize the authors.

For a generalist journal, more could be done to make this study clear, and in particular, to connect to the greater community of proteostasis researchers. I think this study needs a schematic diagram that outlines exactly what was accomplished here, at the beginning. Diagrams like this are especially important for studies like this one that offer a clear and direct set of findings, but conduct many different sorts of tests to get there. I recommend developing a visual abstract that would orient the readers to the work that has been done.

Next, I will make some more specific suggestions. In general, this study is well done and rigorous, but doesn't adequately address a growing literature that examines how proteostasis machinery influences molecular evolution in bacteria.

While this paper might properly test the authors' claims about protein quality control and evolution, the paper does not engage a growing literature in this arena and is generally not very strong on the use of evolutionary theory. I recognize that this is not the aim of the paper, however, and I do not question the authors' authority on the topic. My thoughts here are less about the invocation of theory in evolution (which can be verbose and not relevant), and more about engagement with a growing literature in this very area.

The authors mention Rodrigues 2016, but there are many other studies that should be engaged when discussing the interaction between protein quality control and evolution.

A 2015 study demonstrated how proteostasis machinery can act as a barrier to the usage of novel genes: Bershtein, S., Serohijos, A. W., Bhattacharyya, S., Manhart, M., Choi, J. M., Mu, W., ... & Shakhnovich, E. I. (2015). Protein homeostasis imposes a barrier to functional integration of horizontally transferred genes in bacteria. PLoS genetics, 11(10), e1005612

A 2019 study examined how Lon deletion influenced resistance mutations in DHFR specifically: Guerrero RF, Scarpino SV, Rodrigues JV, Hartl DL, Ogbunugafor CB. The proteostasis environment shapes higher-order epistasis operating on antibiotic resistance. Genetics. 2019 Jun 1;212(2):565-75.

A 2020 study did something similar: Thompson, Samuel, et al. "Altered expression of a quality control protease in E. coli reshapes the in vivo mutational landscape of a model enzyme." Elife 9 (2020): e53476.

And there's a new review (preprint) on this very topic that speaks directly to the various ways proteostasis shapes molecular evolution:

Arenas, Carolina Diaz, Maristella Alvarez, Robert H. Wilson, Eugene I. Shakhnovich, C. Brandon Ogbunugafor, and C. Brandon Ogbunugafor. "Proteostasis is a master modulator of molecular evolution in bacteria."

I am not simply attempting to list studies that should be cited, but rather, this study needs to be better situated in the contemporary discussion on how protein quality control is shaping evolution. This study adds to this list and is a unique and important contribution. However, the findings can be better summarized within the context of the current state of the field. This should be relatively easy to implement.

Reviewer #3 (Public review):

Summary:

This paper investigates the relationship between the proteolytic stability of an antibiotic target enzyme and the evolution of antibiotic resistance via increased gene copy number. The target of the antibiotic trimethoprim is dihydrofolate reductase (DHFR). In Escherichia coli, DHFR is encoded by folA and the major proteolysis housekeeping protease is Lon (lon). In this manuscript, the authors report the results of the experimental evolution of a lon mutant strain of E. coli in response to sub-inhibitory concentrations of the antibiotic trimethoprim and then investigate the relationship between proteolytic stability of DHFR mutants and the evolution of folA gene duplication. After 25 generations of serial passaging in a fixed concentration of trimethoprim, the authors found that folA duplication events were more common during the evolution of the lon strain, than the wt strain. However, with continued passaging, some folA duplications were replaced by a single copy of folA containing a trimethoprim resistance-conferring point mutation. Interestingly, the evolution of the lon strain in the setting of increasing concentrations of trimethoprim resulted in evolved strains with different levels of DHFR expression. In particular, some strains maintained two copies of a mutant folA that encoded an unstable DHFR. In a lon+ background, this mutant folA did not express well and did not confer trimethoprim resistance. However, in the lon- background, it displayed higher expression and conferred high-level trimethoprim resistance. The authors concluded that maintenance of the gene duplication event (and the absence of Lon) compensated for the proteolytic instability of this mutant DHFR. In summary, they provide evidence that the proteolytic stability of an antibiotic target protein is an important determinant of the evolution of target gene copy number in the setting of antibiotic selection.

Strengths:

The major strength of this paper is identifying an example of antibiotic resistance evolution that illustrates the interplay between the proteolytic stability and copy number of an antibiotic target in the setting of antibiotic selection. If the weaknesses are addressed, then this paper will be of interest to microbiologists who study the evolution of antibiotic resistance.

Weaknesses:

Although the proposed mechanism is highly plausible and consistent with the data presented, the analysis of the experiments supporting the claim is incomplete and requires more rigor and reproducibility. The impact of this finding is somewhat limited given that it is a single example that occurred in a lon strain and compensatory mutations for evolved antibiotic resistance mechanisms are described. In this case, it is not clear that there is a functional difference between the evolution of copy number versus any other mechanism that meets a requirement for increased "expression demand" (e.g. promoter mutations that increase expression and protein stabilizing mutations).

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary:

The study by Jena et al. addresses important questions on the fundamental mechanisms of genetic adaptation, specifically, does adaptation proceed via changes of copy number (gene duplication and amplification "GDA") or by point mutation. While this question has been worked on (for example by Tomanek and Guet) the authors add several important aspects relating to resistance against antibiotics and they clarify the ability of Lon protease to reduce duplication formation (previous work was more indirect).

A key finding Jena et al. present is that point mutations after significant competition displace GDA. A second one is that alternative GDA constantly arise and displace each other (see work on GDA-2 in Figure 3). Finally, the authors found epistasis between resistance alleles that was contingent on lon. Together this shows an intricate interplay of lon proteolysis for the evolution and maintenance of antibiotic resistance by gene duplication.

Strengths:

The study has several important strengths: (i) the work on GDA stability and competition of GDA with point mutations is a very promising area of research and the authors contribute new aspects to it, (ii) rigorous experimentation, (iii) very clearly written introduction and discussion sections. To me, the best part of the data is that deletion of lon stimulates GDA, which has not been shown with such clarity until now.

Weaknesses:

The minor weaknesses of the manuscript are a lack of clarity in parts of the results section (Point 1) and the methods (Point 2).

We thank the reviewer for their comments and suggestions on our manuscript. We also appreciate the succinct summary of primary findings that the Reviewer has taken cognisance of in their assessment, in particular the association of the Lon protease with the propensity for GDAs as well as its impact on their eventual fate. We have now revised the manuscript for greater clarity as suggested by Reviewer #1.

Reviewer #2 (Public review):

Summary:

In this strong study, the authors provide robust evidence for the role of proteostasis genes in the evolution of antimicrobial resistance, and moreover, for stabilizing the proteome in light of gene duplication events.

Strengths:

This strong study offers an important interaction between findings involving GDA, proteostasis, experimental evolution, protein evolution, and antimicrobial resistance. Overall, I found the study to be relatively well-grounded in each of these literatures, with experiments that spoke to potential concerns from each arena. For example, the literature on proteostasis and evolution is a growing one that includes organisms (even micro-organisms) of various sorts. One of my initial concerns involved whether the authors properly tested the mechanistic bases for the rule of Lon in promoting duplication events. The authors assuaged my concern with a set of assays (Figure 8).

More broadly, the study does a nice job of demonstrating the agility of molecular evolution, with responsible explanations for the findings: gene duplications are a quick-fix, but can be out-competed relative to their mutational counterparts. Without Lon protease to keep the proteome stable, the cell allows for less stable solutions to the problem of antibiotic resistance.

The study does what any bold and ambitious study should: it contains large claims and uses multiple sorts of evidence to test those claims.

Weaknesses:

While the general argument and conclusion are clear, this paper is written for a bacterial genetics audience that is familiar with the manner of bacterial experimental evolution. From the language to the visuals, the paper is written in a boutique fashion. The figures are even difficult for me - someone very familiar with proteostasis - to understand. I don't know if this is the fault of the authors or the modern culture of publishing (where figures are increasingly packed with information and hard to decipher), but I found the figures hard to follow with the captions. But let me also consider that the problem might be mine, and so I do not want to unfairly criticize the authors.

For a generalist journal, more could be done to make this study clear, and in particular, to connect to the greater community of proteostasis researchers. I think this study needs a schematic diagram that outlines exactly what was accomplished here, at the beginning. Diagrams like this are especially important for studies like this one that offer a clear and direct set of findings, but conduct many different sorts of tests to get there. I recommend developing a visual abstract that would orient the readers to the work that has been done.

The reviewer’s comments regarding data presentation are well-taken. Since we already had a diagrammatic model that sums up the chief findings of our study (Figure 9), we have now provided schematics in Figures 1, 3, 5 and 8 to clarify the workflow of smaller sections of the study. We hope that these diagrams provide greater clarity with regards to the experiments we have conducted.

Next, I will make some more specific suggestions. In general, this study is well done and rigorous, but doesn't adequately address a growing literature that examines how proteostasis machinery influences molecular evolution in bacteria.

While this paper might properly test the authors' claims about protein quality control and evolution, the paper does not engage a growing literature in this arena and is generally not very strong on the use of evolutionary theory. I recognize that this is not the aim of the paper, however, and I do not question the authors' authority on the topic. My thoughts here are less about the invocation of theory in evolution (which can be verbose and not relevant), and more about engagement with a growing literature in this very area.

The authors mention Rodrigues 2016, but there are many other studies that should be engaged when discussing the interaction between protein quality control and evolution.

A 2015 study demonstrated how proteostasis machinery can act as a barrier to the usage of novel genes: Bershtein, S., Serohijos, A. W., Bhattacharyya, S., Manhart, M., Choi, J. M., Mu, W., ... & Shakhnovich, E. I. (2015). Protein homeostasis imposes a barrier to functional integration of horizontally transferred genes in bacteria. PLoS genetics, 11(10), e1005612

A 2019 study examined how Lon deletion influenced resistance mutations in DHFR specifically: Guerrero RF, Scarpino SV, Rodrigues JV, Hartl DL, Ogbunugafor CB. The proteostasis environment shapes higher-order epistasis operating on antibiotic resistance. Genetics. 2019 Jun 1;212(2):565-75.

A 2020 study did something similar: Thompson, Samuel, et al. "Altered expression of a quality control protease in E. coli reshapes the in vivo mutational landscape of a model enzyme." Elife 9 (2020): e53476.

And there's a new review (preprint) on this very topic that speaks directly to the various ways proteostasis shapes molecular evolution:

Arenas, Carolina Diaz, Maristella Alvarez, Robert H. Wilson, Eugene I. Shakhnovich, C. Brandon Ogbunugafor, and C. Brandon Ogbunugafor. "Proteostasis is a master modulator of molecular evolution in bacteria."

I am not simply attempting to list studies that should be cited, but rather, this study needs to be better situated in the contemporary discussion on how protein quality control is shaping evolution. This study adds to this list and is a unique and important contribution. However, the findings can be better summarized within the context of the current state of the field. This should be relatively easy to implement.

We thank the reviewer for their encouraging assessment of our manuscript as well as this important critique regarding the context of other published work that relates proteostasis and molecular evolution. Indeed, this was a particularly difficult aspect for us given the different kinds of literature that were needed to make sense of our study. We have now added the references suggested by the reviewer as well as others to the manuscript. We have also added a paragraph in the discussion section (Lines 463-476) that address this aspect and hopefully fill the lacuna that the reviewer points out in this comment.

Reviewer #3 (Public review):

Summary:

This paper investigates the relationship between the proteolytic stability of an antibiotic target enzyme and the evolution of antibiotic resistance via increased gene copy number. The target of the antibiotic trimethoprim is dihydrofolate reductase (DHFR). In Escherichia coli, DHFR is encoded by folA and the major proteolysis housekeeping protease is Lon (lon). In this manuscript, the authors report the results of the experimental evolution of a lon mutant strain of E. coli in response to sub-inhibitory concentrations of the antibiotic trimethoprim and then investigate the relationship between proteolytic stability of DHFR mutants and the evolution of folA gene duplication. After 25 generations of serial passaging in a fixed concentration of trimethoprim, the authors found that folA duplication events were more common during the evolution of the lon strain, than the wt strain. However, with continued passaging, some folA duplications were replaced by a single copy of folA containing a trimethoprim resistance-conferring point mutation. Interestingly, the evolution of the lon strain in the setting of increasing concentrations of trimethoprim resulted in evolved strains with different levels of DHFR expression. In particular, some strains maintained two copies of a mutant folA that encoded an unstable DHFR. In a lon+ background, this mutant folA did not express well and did not confer trimethoprim resistance. However, in the lon- background, it displayed higher expression and conferred high-level trimethoprim resistance. The authors concluded that maintenance of the gene duplication event (and the absence of Lon) compensated for the proteolytic instability of this mutant DHFR. In summary, they provide evidence that the proteolytic stability of an antibiotic target protein is an important determinant of the evolution of target gene copy number in the setting of antibiotic selection.

Strengths:

The major strength of this paper is identifying an example of antibiotic resistance evolution that illustrates the interplay between the proteolytic stability and copy number of an antibiotic target in the setting of antibiotic selection. If the weaknesses are addressed, then this paper will be of interest to microbiologists who study the evolution of antibiotic resistance.

Weaknesses:

Although the proposed mechanism is highly plausible and consistent with the data presented, the analysis of the experiments supporting the claim is incomplete and requires more rigor and reproducibility. The impact of this finding is somewhat limited given that it is a single example that occurred in a lon strain and compensatory mutations for evolved antibiotic resistance mechanisms are described. In this case, it is not clear that there is a functional difference between the evolution of copy number versus any other mechanism that meets a requirement for increased "expression demand" (e.g. promoter mutations that increase expression and protein stabilizing mutations).

We thank the reviewer for their in-depth assessment of our work and appreciate their concerns regarding reproducibility and rigor in analysis of our data. We have now incorporated this feedback and provided necessary clarifications/corrections in the revised version of our manuscript.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

Major Points:

(1) The authors show that a deletion of lon increases the ability for GDA and they argue that this is adaptive during TMP treatment because it increases the dosage of folA (L. 129). However, the highest frequency of GDA occurred in drug-free conditions (see Figure 1C). This indicates either that GDA is selected in drug-free media and potentially selected against by certain antibiotics. It would help for the authors to discuss this possibility more clearly.

We thank the reviewer for this astute observation. It is indeed striking that the GDA mutation (i.e. the GDA-2 mutation) selected in a lon-deficient background does not come up in presence of antibiotics. To probe this further, we have now measured the relative fitness of a representative population of lon-knockout from short-term evolution in drug-free LB (population #3) that harbours GDA-2 against its ancestor (marked with DlacZ). These competition experiments were performed in LB (in which GDA-2 emerged spontaneously), as well as in LB supplemented with antibiotics at the concentrations used during the short term evolution.

Values of relative fitness, w (mean ± SD from 3 measurements), are provided below:

LB: 1.4 ± 0.2

LB + Trimethoprim: 1.6 ± 0.2

LB + Spectinomycin: 0.9 ± 0.2

LB + Erythromycin: 1.3 ± 0.3

LB + Nalidixic acid: 1.5 ± 0.2

LB + Rifampicin: 1.4 ± 0.2

These data show an increase in relative fitness in drug-free LB as would be expected. Interestingly, we also observe an increase in relative fitness in LB supplemented with antibiotics, except spectinomycin. This result supports the idea that GDA-2 is a “media adaptation” and provides a general fitness advantage to the lon knockout. However, as the reviewer pointed out, we should expect to see GDA-2 emerge spontaneously in antibiotic-supplemented media as well. We think that this does not happen as the fitness advantage of drug-specific mutations (GDAs or point mutations) far exceed the advantage of a media adaptation GDA. As a result, we only see the specific mutations that provide high benefit against the antibiotic at least over the relatively short duration of 20-25 generations. It is noteworthy the GDA-2 mutation does come up in LTMPR1 when it is passaged over >200 generations in drug-free media, but shows fluctuating frequency over time. We expect, therefore, that given enough time we may detect the GDA-2 mutations even in antibiotic-supplemented media.

We note, however, that a major caveat in the above fitness calculations is that we cannot be sure that the competing ancestor has no GDA-2 mutations during the course of the experiment. Thus, the above fitness values are only indicative and not definitive. We have therefore not included these data in the revised manuscript.

(2) It is unclear if the isolates WTMPR1 - 5 and LTMPR1 - 5 were pure clones. The authors write in L.488 "Colonies were randomly picked, cultured overnight in drug-free LB and frozen in 50% glycerol at -80C until further use." And in L. 492 "For long-term evolution, trimethoprim-resistant isolates LTMPR1, WTMPR4 and WTMPR5 were first revived from frozen stocks in drug-free LB overnight." From these descriptions, it is possible that the isolates contained a fraction of cells of other genotypes since colonies are often formed by more than one cell and thus, unless pure-streaked, a subpopulation is present and would in drug-free media be maintained. The possibility of pre-existing subpopulations is important for all statements relating to "reversal".

This is indeed a valid concern. As far as we can tell all our initial isolates (i.e. WTMPR1-5 and LTMPR1-5) are pure clones at least as far as SNPs are concerned. This is based on whole genome sequencing data that we have reported earlier in Patel and Matange, eLife (2021), where we described the evolution and isolation of WTMPR1-5 and the present study for LTMPR1-5. All SNPs detected were present at a frequency of 100%. For clones with GDAs, however, there is no way to eliminate a sub-population that has a lower or higher gene copy number than average from an isolate. This is because of the inherent instability of GDAs that will inevitably result in heterogeneous gene copy number during standard growth. In this sense, there is most certainly a possibility of a pre-existing subpopulation within each of the clones that may have reversed the GDA. Indeed, we believe that it is this inherent instability that contributes to their rapid loss during growth in drug-free media.

Minor Points:

(1) L. 406. "allowing accumulation of IS transposases in E. coli" Please specify that it is the accumulation of transposase proteins (and not genes).

We have made this change.

(2) L. 221 typo. Known "to" stabilize.

We have made this change.

Reviewer #2 (Recommendations for the authors):

Most of my suggestions are found in the public review. I believe this to be a strong study, and some slight fixes can solidify its presence in the literature.

We have attempted to address the two main critiques by Reviewer 2. To simplify the understanding of our data, we have provided small schematics at various points in the paper to clarify the experimental pipelines used by us. We have also provided additional discussion situating our study in the emerging area of proteostasis and molecular evolution. We hope that our revisions have addressed these lacunae in our manuscript.

Reviewer #3 (Recommendations for the authors):

Major Points:

(1) The manuscript is generally a bit difficult to follow. The writing is overly complicated and lacks clarity at times. It should be simplified and improved.

We have made several revisions to the text, as well as provided schematics in some of our figures which hopefully make our paper easier to understand.

(2) I cannot find the raw variant summary data for the lon strain evolution experiment in trimethoprim (after 25 generations). Were there any other mutations identified? If not, this should be explicitly stated in the text and the variant output summary from sequencing included as supplemental data.

We apologise for this oversight. We have now provided these data as Table 1.

(3) What is the trimethoprim IC50 of the starting (pre-evolution) strains (i.e. wt and lon)? I can't find this information, but it is critical to interpretation.

We had reported these values earlier in Matange N., J Bact (2020). Wild type and lon-knockout have similar MIC values for trimethoprim, though the lon mutant shows a higher IC50 value. We have now mentioned this in the results section (Line 100-101) and also provided the reference for these data.

(4) What was the average depth of coverage for WGS? This information is necessary to assess the quality of the variant calling, especially for the population WGS.

All genome sequencing data has a coverage at least 100x. We have added this detail to the methods section (Line 580-581).

(5) Five replicate evolution experiments (25 generations, or 7x 10% daily batch transfers) were performed in trimethoprim for the wt and lon strains. Duplication of the folA locus occurred in 1/5 and 4/5 experiments, respectively. It is not entirely clear what type of sampling was actually done to arrive at these numbers (this needs to be stated more clearly), but presumably 1 random colony was chosen at the end of the passaging protocol for each replicate. Based on this result, the authors conclude that folA duplication occurred more frequently in the lon strain, however, this is not rigorously supported by a statistical evaluation. With N=5, one cannot rigorously conclude that a 20% frequency and 80% frequency are significantly different. Furthermore, it's not entirely clear what the mechanism of resistance is for these strains. For example, in one colony sequenced (LTMPR5), it appears no known resistance mechanism (or mutations?) were identified, and yet the IC50 = 900 nM, which is also similar to other strains.

Indeed, we agree with the reviewer that we don’t have the statistical power to rigorously make this claim. However, since the lon-knockout showed us a greater frequency of GDA across 3 different environments we are fairly confident that loss of lon enhances the overall frequency for GDA mutations. This idea in also supported by a number of previous papers that related GDAs and IS-element transpositions with Lon, viz. Nicoloff et al, Antimicrob Agent Chemother (2007), Derbyshire et al. PNAS (1990), Derbyshire and Grindley, Mol Microbiol (1996). We have therefore not provided further justification in the revised manuscript.

We had indeed sampled a random isolate from each of the 5 populations and have added a schematic to figure 1 that provides greater clarity.

Having relooked at the sequencing data for LTMPR1-5 isolates (Table 1), we realised that both LTMPR4 and LTMPR5 harbour mutations in the pitA gene. We had missed this locus during the previous iteration of this manuscript and misidentified an mgrB mutations in LTMPR4. PitA codes for a metal-phosphate symporter. We have observed mutations in pitA in earlier evolution experiments with trimethoprim as well (Vinchhi and Yelpure et al. mBio 2023). Interestingly, in LTMPR5 there was a deletion of pitA, along with 17 other contiguous genes mediated by IS5. To test if loss of pitA is beneficial in trimethoprim, we tested the ability of a pitA knockout to grow on trimethoprim supplemented plates. Indeed, loss of pitA conferred a growth advantage to E. coli on trimethoprim, comparable to loss of mgrB, indicating that the mechanism of resistance of LTMPR5 may be due to loss of pitA. We have added these data to the Supplementary Figure 1 of the revised manuscript and provided a brief description in Lines 103-108. How pitA deficiency confers trimethoprim resistance is yet to be investigated. The mechanism is likely to be by activating some intrinsic resistance mechanism as loss of pitA also conferred a fitness benefit against other antibiotics. This work is currently underway in our lab and hence we do not provide any further mechanism in the present manuscript.

(6) Although measurement error/variance is reported, statistical tests were not performed for any of the experiments. This is critical to support the rigor and reproducibility of the conclusions.

We have added statistical testing wherever appropriate to the revised manuscript.

(7) Lines 150-155 and Figure 2E: Putting a wt copy of mgrB back into the WTMPR4 and LTMPR1 strains would be a better experiment to dissect out the role of mgrB versus the other gene duplications in these strains on fitness. Without this experiment, you cannot confidently attribute the fitness costs of these strains to the inactivation of mgrB alone.

We agree with the reviewer that our claim was based on a correlation alone. We have now added some new data to confirm our model (Figure 2 E, F). The costs of mgrB mutations come from hyperactivation of PhoQP. In earlier work we have shown that the costs (and benefit) of mgrB mutations can be abrogated in media supplemented with Mg2+, which turns off the PhoQ receptor (Vinchhi and Yelpure et al. mBio, 2023). We use this strategy to show that like the mgrB-knockout, the costs of WTMPR4, WTMPR5 and LTMPR1 can be almost completely alleviated by adding Mg2+ to growth media. These results confirm that the source of fitness cost of TMP-resistant bacteria was not linked to GDA mutations, but to hyperactivation of PhoQP.

(8) Figure 3F and G: Does the top symbol refer to the starting strain for the 'long-term' evolution? If so, why does WTMPR4 not have the mgrB mutation (it does in Figure 1)? Based on your prior findings, it seems odd that this strain would evolve an mgrB loss of function mutation in the absence of trimethoprim exposure.

We thank the reviewer for pointing this error out. We have made the correction in the revised manuscript.

(9) Figure 6A: If the marker is neutral, it should be maintained at 0.1% throughout the 'neutrality' experiment. In both plots, the proportion of some marked strains goes up and then down. This suggests either ongoing evolution (these competitions take place over 105 generations), or noisy data. I suspect these data are just inherently noisy. I don't see error bars in the plots. Were these experiments ever replicated? It seems that replicating the experiments might be able to separate out noise from signal and perhaps clarify this point and better confirm the hypothesis that the point mutants are more fit.

These experiments were indeed noisy and the apparent enrichment is most likely a measurement error rather than a real change in frequency of competing genotypes. We have now provided individual traces for each of the competing pairs with mean and SD from triplicate observations at each time point.

(10) Figure 6A: Please indicate which plotted line refers to which 'point mutant' using different colors. These mutants have different trimethoprim IC50s and doubling times, so it would be nice to be able to connect each mutant to its specific data plot.

We thank the reviewer for this suggestion. We have now colour coded the different strain combinations as suggested.

(11) Lines 284-285: I disagree that the IC50s are similar. The C-35T mutant has IC50 that is 2x that of LTMPR1. Perhaps more telling is that, compared to the folA duplication strain from the same time-point (which also carries the rpoS mutation), all of the point mutants have greater IC50s (~2x greater). 2-fold changes in IC50 are significant. It would seem that the point-mutants were likely not competing against LTMPR1 at the time they arose, so LTMPR1 might not be the best comparator if it was extinguished from the population early. I'm assuming this is why you chose a contemporary isolate (and, also, rpoS mutant) for the competition experiments. This should be explained more clearly.

We thank the reviewer for this comment. Indeed, the reviewer is correct about the rationale behind the use of a contemporary isolate and we have provided this clarification in the revised manuscript (Line 287-289). Also, the reviewer is correct in pointing out that a two-fold difference in IC50 cannot be ignored. However, the key point here would be in assessing the differences in growth rates at the antibiotic concentration used during competition (i.e. 300 ng/mL). We are unable to see a direct correlation between the growth rates and enrichment in culture indicating that the observed trends are unlikely to be driven by ‘level of resistance’ alone. We have added these clarifications to the modified manuscript (Lines 299-301)

Minor Points:

(1) Line 13: Add a comma before 'Escherichia'

We have made this change.

(2) Line 14: Consider changing "mutations...were beneficial in trimethoprim" to "mutations...were beneficial under trimethoprim exposure"

We have made this change.

(3) Line 32: Is gene dosage really only "relative to the genome"? Is it not simply its relative copy number generally? Consider changing to "The dosage of a gene, or its relative copy number, can impact its level of expression..."

We have made this change.

(4) Line 38: The idea that GDAs are 1000x more frequent than point mutations seems an overgeneralization.

We agree with the reviewer and have softened our claim.

(5) Line 50: The term "hard-wired" is confusing. Please be more specific.

We have modified this statement to “…GDAs are less stable than point mutations….”.

(6) Line 52-53: What do you mean by "there is also evidence to suggest that...more common in bacteria than appreciated"? Are you implying the field is naïve to this fact? If there is "evidence" of this, then a reference should be included. However, it's not clear why this is important to state in the article. I would consider simply removing this sentence. Less is more in this case.

We have removed this statement.

(7) Lines 59-60: Enzymes catalyze reactions. Please also state the substrates for DHFR. Consider, "It catalyzes the NADPH-dependent reduction of dihydrofolate to tetrahydrofolate, and important co-factor for..."

We have made this change.

(8) Line 72: Please change to, "In E. coli, DHFR is encoded by folA." You do not need to state this is a gene, as it is implicit with lowercase italics.

We have made this change.

(9) Lines 72-86: This paragraph is a bit confusing to read, as it has several different ideas in it. Consider breaking it into two paragraphs at Line 80, "In this study,...". The first paragraph could just review the trimethoprim resistance mechanisms in E. coli and so would change the first sentence (Line 72) to reflect this topic: "In E. coli, DHFR is encoded by folA and several different resistance mechanisms have been characterized." Then, just describe each mechanism in turn. Also, by "hot spots" it would seem you are referring to "point mutations" in the gene that alter the protein sequence and cluster onto the 3D protein structure when mapped? Please be more specific with this sentence for clarity.

We have made these changes.

(10) Lines 92-93: Please also state the MIC value of the strain to specifically define "sub-MIC". Alternatively, you could also state the fraction MIC (e.g. 0.1 x MIC).

We have modified this statement to “…in 300 ng/mL of trimethoprim (corresponding to ~0.3 x MIC) for 25 generations.”

(11) Lines 95-96. Remove, "These sequencing have been reported earlier, ...(2021)". You just need to cite the reference.

We have made this change.

(12) Line 96: Remove the word "gene".

We have made this change.

(13) Figure 1 and Figure 4C: The color scheme is tough for those with the most common type of color blindness. Red/green color deficiency causes a lot of difficulty with Red/gray, red/green, green/gray. Consider changing.

We thank the reviewer for bringing this to our notice. We have modified the colour scheme throughout the manuscript.

(14) Figure 1: Was there a trimethoprim resistance mechanism identified for LTMPR5?

As stated by us in response to major comment #7, LTMPR5’s resistance seems to come from a novel mechanism involving loss of the pitA gene.

(15) Line 349-351: Please briefly define "lower proteolytic stability" as a relative susceptibility to proteolytic degradation and make sure it is clear to the reader that this causes less DHFR. This needs to be clarified because it is confusing how a mutation that causes DHFR proteolytic instability would lead to an increase in trimethoprim IC50. So, you also need to mention that some mutations can cause both increased trimethoprim inhibition and lower proteolytic stability simultaneously. It seems the Trp30Arg mutation is an example of this, as this mutation is associated with a net increase in trimethoprim resistance despite the competing effects of the mutation on enzyme inhibition and DHFR levels.

We thank the reviewer for this comment and agree that the text in the original manuscript did not fully convey the message. We have made modifications to this section (Lines 359-363) in the revised manuscript in agreement with the reviewer’s suggestions.

Version 2

Abstract

Abstract

Evolution of gene expression frequently drives antibiotic resistance in bacteria. We had previously (Patel and Matange, eLife, 2021) shown that, in Escherichia coli, mutations at the mgrB locus were beneficial under trimethoprim exposure and led to overexpression of dihydrofolate reductase (DHFR), encoded by the folA gene. Here, we show that DHFR levels are further enhanced by spontaneous duplication of a genomic segment encompassing folA and spanning hundreds of kilobases. This duplication was rare in wild type E. coli. However, its frequency was elevated in a lon-knockout strain, altering the mutational landscape early during trimethoprim adaptation. We then exploit this system to investigate the relationship between trimethoprim pressure and folA copy number. During long-term evolution, folA duplications were frequently reversed. Reversal was slower under antibiotic pressure, first requiring the acquisition of point mutations in DHFR or its promoter. Unexpectedly, despite resistance-conferring point mutations, some populations under high trimethoprim pressure maintained folA duplication to compensate for low abundance DHFR mutants. We find that evolution of gene dosage depends on expression demand, which is generated by antibiotic and exacerbated by proteolysis of drug-resistant mutants of DHFR. We propose a novel role for proteostasis as a determinant of copy number evolution in antibiotic-resistant bacteria.

Introduction

The dosage of a gene, i.e., relative copy number, directly impacts its level of expression and can alter organismal phenotype and fitness [1–3]. Mutations that increase copy numbers, termed as gene duplications and amplifications (GDAs), are implicated in diverse biological phenomena including cancer progression [4–6], speciation [7–9], reproductive isolation [10–12] and antimicrobial resistance (AMR) [2, 13]. GDAs are important for rapid adaptation since their frequencies of occurrence can be higher by several orders of magnitude than point mutations [2, 13–16]. Following the initial GDA mutation, however, amplified genes could be lost over time, pseudogenized, sub-functionalized or neo-functionalized [17–19]. Which of these possibilities is realized depends on the nature and strength of selection imposed by the environment, as well as the fitness cost associated with maintaining multiple gene copies [17–19].

In bacteria, GDA mutations are frequently encountered during adaptation to antibiotics [13] or heavy metals [20, 21]. They occur by RecA-dependent nonequal recombination as well as RecA-independent processes and often involve repeat sequences such as in IS-elements [22–25]. Amplified genes in drug-resistant bacteria code for effectors of resistance such as drug inactivating enzymes, drug target modifiers or “by-pass” mechanisms [26–28], but may also compensate for resistance-associated fitness costs [29]. More recently, the view that GDAs are less stable than point mutations has emerged. For instance, transient alterations in gene dosage are a common mechanism of heteroresistance, i.e. resistant sub-populations within a drug-sensitive bacterial isolate [20, 30, 31]. Further, it is estimated that as many as 20% of cells in a bacterial population may harbour a duplication of some genomic region [32]. These copy number polymorphisms create within-population heterogeneities in gene expression, serving as a bet-hedging strategy with adaptive value in fluctuating environments [33].

Dihydrofolate reductase (DHFR) is a ubiquitous enzyme crucial for growth and division. It catalyzes the NADPH-dependent reduction of tetrahydrofolate, an important cofactor in nucleotide and amino acid biosynthesis. DHFR is coded by an essential gene in most organisms and its deletion or pharmacological inhibition results in cell death or growth stasis [34]. Not surprisingly, anti-folates are effective therapeutic agents for diverse pathologies [35, 36]. For instance, methotrexate, used in the treatment of cancers and autoimmune conditions [37], pyrimethamine, used as an anti-malarial [38], and trimethoprim, used as an anti-bacterial [39], are all competitive inhibitors of DHFR. Interestingly, copy number amplification and consequent overexpression of DHFR confers resistance in all three systems [40–43]. Thus, the gene dosage of DHFR appears to be tightly regulated across evolutionarily distant organisms, making it a useful system to understand the principles underlying selection and evolutionary fate of GDA mutations.

In Escherichia coli DHFR is coded by folA. Several point mutations that confer trimethoprim resistance have been identified in this enzyme [44–47]. Additionally, mechanisms that transcriptionally up-regulate folA expression are also known in trimethoprim resistant E. coli. These include cis-regulatory mutations in the folA promoter and trans-acting loss of function mutations in mgrB, which upregulate folA transcription in a PhoQP-dependent manner [44–46, 48]. Earlier work from our group identified Lon, a conserved quality control protease in bacteria, as a post-translational regulator of trimethoprim resistance. Lon degraded mutant DHFRs and reduced their half-lives in vivo, directly affecting genotype-phenotype relationships of resistance-conferring mutations [49, 50].

In this study, we first compare early mutational events in wild type and Lon-deficient E. coli challenged with trimethoprim. We show that a large genomic duplication encompassing folA confers trimethoprim resistance in both genetic backgrounds, though its frequency of occurrence is elevated in lon-knockout E. coli. We then use this system to ask what factors determine the fate of duplicated folA gene copies during evolution at different trimethoprim pressures.

Results

A large IS-flanked duplication encompassing folA mediates rapid adaptation to trimethoprim in E. coli

We had previously evolved 5 replicate populations of wild type E. coli K-12 MG1655 in 300 ng/mL of trimethoprim ( ∼0.3 x MIC) for 25 generations. We then isolated and sequenced the genome of a dominant trimethoprim-resistant clone from each population (designated as WTMPR1-5) (Figure 1 A). These sequencing results have been reported earlier [46]. Briefly, all isolates harboured loss-of-function mutations in the mgrB gene (Figure 1 B) that resulted in PhoQP-dependent overexpression of folA [46]. Additionally, isolates WTMPR1, 3 and 5 had mutations at the active site of DHFR, while WTMPR4 harboured a large genomic duplication of ∼0.6 Mb encompassing folA (Figure 1 B, C). We performed similar evolution experiments starting with an isogenic E. coli Δlon strain to ask if early adaptation to trimethoprim was altered by Lon-deficiency (Figure 1 A). Wild type and lon-deficient E. coli have similar MIC for trimethoprim, though the lon-knockout has a higher IC50 [49]. Among trimethoprim-resistant isolates derived from E. coli Δlon, designated as LTMPR1-5, three harboured mutations at the mgrB locus (LTMPR1-3, Figure 1 A, B, Table 1). LTMPR4 and LTMPR5 harboured mutations in pitA, coding for a metal-phosphate symporter, that have been reported earlier in trimethoprim adapted E. coli [51]. LTMPR5 had a deletion encompassing pitA, indicating that loss of PitA was beneficial in trimethoprim. A knockout of pitA indeed showed greater colony formation on trimethoprim-supplemented media demonstrating that mutations in this gene were adaptive (Figure S1), though the underlying molecular mechanisms are unknown at present. Interestingly, none of the LTMPR isolates had mutations in DHFR or its promoter. Instead, LTMPR2-4 had evolved GDA mutations similar to WTMPR4, while LTMPR1 had a duplication in an overlapping but shorter genomic stretch (Figure 1 B, C, Table 1).

Adaptation to antibiotics by large genomic duplications is favoured in lon- deficient E. coli.

A. Schematic of the experimental pipeline used to compare the mutation repertoire of adaptation to trimethoprim in wild type or lon-deficient E. coli. B. Coverage depth plots and mutations at the folA and mgrB loci of 5 independently evolved trimethoprim-resistant isolates derived from wild type (WTMPR1-5) or Δlon E. coli (LTMPR1-5). Coverage depth plots show the number of reads from Illumina short-read sequencing (y-axis) mapped to each genomic coordinate of the reference genome (x-axis). Coverage of 1x, 2x and 3x are marked by dotted lines for reference. Mutations in mgrB and folA, and trimethoprim IC50 values (mean ± S.D. from 3 independent measurements) are provided for each isolate below the appropriate coverage depth plot. C. Cartoon (not to scale) showing the duplicated genomic stretch and flanking IS-elements in trimethoprim-resistant isolates. Blue arrows represent IS1, while red arrows represent IS186. The positions of folA and acrAB genes, implicated in trimethoprim resistance in earlier studies, relative to the duplicated stretch are shown. Fold increase in the number of reads corresponding to the folA gene are indicated for each isolate. D. Summary of the prevalence of GDA mutations detected in wild type or E. coli Δlon after ∼25 generations of evolution in control media (no antibiotic) or in media supplemented with sub-MIC antibiotics. Each circle represents one of the 5 replicates in the evolution experiment. Empty circles represent replicates in which no resistance was detected (dotted perimeter) or where the whole population or resistant isolates were sequenced but no GDA mutation was detected (solid perimeter). Red filled circles indicate that a GDA mutation was identified.

Duplicated genomic regions in WTMPR4 and LTMPR1-4 were flanked by IS-elements (Figure 1 C). Seeking to explain the higher frequency of GDA mutations in the trimethoprim-adapted lon-knockout, we asked whether IS-transposases were sensitive to proteolysis by Lon. Indeed, heterologously expressed InsB and InsL transposases had greater stability in the lon-knockout (Figure S1), suggesting that their accumulation may contribute to higher GDA frequency in E. coli Δlon. It would follow that a higher proportion of GDA mutants would be isolated from E. coli Δlon during adaptation in other environments as well. In agreement, GDA mutations were detected after evolution of E. coli Δlon in 2 of the 5 additional growth conditions tested (Figure 1 D, S2). In all cases duplications were flanked by IS elements (Figure S2). Further, duplicated genomic regions were similar among isolates derived from the same growth condition, but different across environments (Figure 1 B, S2). On the other hand, wild type E. coli did not show GDA mutations in any of the other tested conditions (Figure 1 D).

The duplicated region in trimethoprim-resistant isolates included genes previously associated with resistance to trimethoprim, namely, folA, and acrA and B, which code for components of the AcrAB-TolC efflux pump (Figure 1 C). In one of the isolates, i.e. LTMPR1, the amplified stretch encompassed folA but not acrAB without affecting trimethoprim IC50 values (Figure 1 B, C). We therefore reasoned that duplication of folA was the main effector of adaptation to trimethoprim. Immunoblotting showed that DHFR protein was indeed overproduced in isolates LTMPR1 and WTMPR4 over and above the increase mediated by inactivation of mgrB alone (Figure 2 A).

Overexpression of DHFR in trimethoprim-resistant isolates with genomic duplication.

A. DHFR protein expression in trimethoprim-resistant isolates WTMPR4 and LTMPR1, compared with respective ancestors and mgrB-knockout strains. DHFR protein was detected by immunoblotting using an anti-DHFR polyclonal antibody. FtsZ was used as loading control. A representative immunoblot from 3 biological replicates is shown. Quantitation was performed by calculating band intensities using image analysis. Fold change values of DHFR levels over controls (wild type or E. coli Δlon, set to 1) is shown as mean ± S.D. from triplicate measurements. Alleles of mgrB and folA in WTMPR4 and LTMPR1 are shown diagrammatically. B. Expression level of duplicated genes in LTMPR1 and WTMPR4, determined by RNA-sequencing, expressed as fold over E. coli ΔlonΔmgrB and E. coli ΔmgrB respectively. Each point represents fold change value for a single gene. Mean of the scatter is shown as a black line, and its value is provided along with standard deviation above the plot. C, D. Correlation between expression levels of genes in LTMPR1 (C) and WTMPR4 (D) with E. coli ΔlonΔmgrB and E. coli ΔmgrB respectively. Gene expression levels were estimated using RNA-sequencing and fold changes were calculated using wild type E. coli as reference for WTMPR4 and E. coli ΔmgrB, and E. coli Δlon as reference for LTMPR1 and E. coli ΔlonΔmgrB. Each point represents log2(fold change) value for a single gene. The resulting scatter was fit to a simple linear regression and the obtained R2 values are provided. E, F. Relative fitness (w) of indicated E. coli strains calculated using competition experiments with E. coli ΔlacZ as reference in the absence (E) or presence (F) of 10 mM MgSO4. Presence of high concentration of Mg2+ alleviates the fitness cost of mgrB mutations by inhibiting the PhoQP system [51]. For strains derived from E. coli Δlon, E. coli ΔlonΔlacZ was used as reference. Neutrality of ΔlacZ was established by competition with the unmarked ancestral strains, i.e. wild type (WT) and E. coli Δlon. No change in relative fitness compared to the reference strain (w = 1) is indicated by a dotted line. Mean ± S.D. from three independent measurements is plotted. Statistical significance was tested using an unpaired t- test and a p-value of < 0.5 was considered statistically significantly different (*). A p-value ≥ 0.5 was considered not significantly different (ns).

In addition to folA, ∼173 and ∼531 other genes were also duplicated in LTMPR1 and WTMPR4, respectively. To understand the global transcriptional burden imposed by this amplification, we analysed the gene expression profiles of WTMPR4 and LTMPR1 using RNA-sequencing (Supplementary File 1). Since both isolates had loss-of-function mutations in mgrB, which cause large changes in global gene expression [51], we also analysed the transcriptomes of E. coli ΔmgrB and E. coli ΔlonΔmgrB strains for comparison (Supplementary File 1). Transcript level of folA was higher in LTMPR1 (2.3-fold) and WTMPR4 (5.7-fold) over respective controls (Supplementary File 1). Other duplicated genes were mildly overexpressed by ∼1.7-fold on average in LTMPR1 (over E. coli ΔlonΔmgrB) and ∼1.3-fold in WTMPR4 (over E. coli ΔmgrB) (Figure 2 B, Supplementary File 1). However, there was large variation in expression level among duplicated genes (Figure 2 B), indicating that increase in gene dosage did not necessarily result in enhanced expression levels. Instead, global gene expression profiles of WTMPR4 and LTMPR1 were correlated with E. coli ΔmgrB and E. coli ΔlonΔmgrB respectively (Figure 2 C, D) and most of the top 50 up/down-regulated genes in WTMPR4 and LTMPR1 could be attributed to loss of mgrB alone (Figure S3). In line with this result, strains with and without genomic amplifications had comparable fitness costs (∼20 %) in drug-free media, similar in magnitude to the cost of mgrB inactivation (Figure 2 E). Further, the fitness cost was alleviated in growth media supplemented with high Mg2+, i.e. by inhibiting the PhoQP pathway, confirming the source of this cost to be loss of mgrB rather than the GDA (Figure 2 F). Thus, we concluded that large GDA mutations encompassing folA led to DHFR overexpression and represented an alternate mechanism for trimethoprim-resistance in E. coli with undetectable additional cost.

Maintenance of folA duplication is linked to antibiotic-generated demand for DHFR expression

We next investigated the evolutionary fate of duplicate folA gene copies by evolving the LTMPR1 isolate, which harboured the shortest duplication, in the absence of antibiotic. We also evolved this isolate at sustained drug pressure, i.e. trimethoprim at the same concentration at which it had originally emerged (300 ng/mL). Using three replicate populations in each condition, we asked how DHFR expression and trimethoprim IC50 values changed over ∼252 generations of evolution and used genome sequencing to identify the underlying mutational basis. As controls, we performed similar evolution experiments with WTMPR4 (which harboured a GDA and a functional Lon protease) and WTMPR5 (which harboured a folA point mutation and a functional Lon protease) (Figure 3 A).

Evolutionary fate of folA-duplication in the absence of drug pressure.

A. Schematic of the experimental pipeline used to investigate the impact of evolution in drug-free media on resistance level, folA copy number and expression level of different trimethoprim resistant E. coli populations. B. Trimethoprim IC50 values of 3 evolving lineages derived from LTMPR1 (A, B, C) over 252 generations of evolution in antibiotic-free medium. Mean ± S.D. from three measurements is plotted at each time point. IC50 values of E. coli Δlon and ΔlonΔmgrB are shown as dotted lines for reference. C. Colony formation of LTMPR1, WTMPR4 and WTMPR5 evolved in antibiotic-free medium at different generations. Fraction of the population capable of forming colonies at 1, 5 and 10 μg/mL trimethoprim was calculated across 3 replicate lines. Mean value from the three lines is plotted. The results of similar experiments performed on ancestors (0 generations) are also provided for reference. D. DHFR expression during evolution in the absence of trimethoprim in LTMPR1 lineages A, B and C, measured by immunoblotting using anti-DHFR polyclonal antibody. FtsZ was used as a loading control. Quantitation was performed by calculating band intensities using image analysis. DHFR expression at each time point was normalized to the ancestor (i.e. 0 generations, set to 1). Mean of three independent measurements is shown below each lane. E. Copy number of the ancestral GDA encompassing folA (GDA(folA)) and GDA-2 at different time points of evolution in the absence of trimethoprim are plotted. For GDA(folA), copy number was determined by dividing number of reads from an Illumina sequencing experiment corresponding to folA by the average number of reads mapping to the rest of the genome. F. Point mutations in folA, rpoS and mgrB, and the “GDA-2” genomic duplication in 6 randomly picked colonies from each of the LTMPR1 lines at 252 generations of evolution are shown schematically using the appropriate symbols. Asterix (*/**) marks the genotypes that were carried forward for further analyses. G, H. Point mutations in folA, mgrB and rpoS in isolates derived from 252 generations of evolution of WTMPR4 (G) and WTMPR5 (H) in no antibiotic. From each of the evolving lines 2 random isolates were picked for genome sequencing. Various point mutants at the 3 gene loci of interest are represented by appropriate symbols as shown in the legend..

Evolution in trimethoprim-free media

In drug-free conditions, trimethoprim IC50 of LTMPR1 rapidly declined and a concomitant loss of colony formation on trimethoprim-supplemented plates was observed (Figure 3 B, C). Evolved WTMPR4, but not WTMPR5, showed similar loss of colony formation on trimethoprim indicating that resistance mediated by GDA mutations was less stable than point mutations regardless of the presence of Lon (Figure 3 C). For LTMPR1, these trends were explained by a decline in DHFR expression levels over evolution and reversion to a single folA copy (Figure 3 D, E, S4). Mutation in mgrB was, however, consistently maintained till the end of the experiment (Supplementary File 2). In agreement with genomics data, trimethoprim IC50s of evolving populations stabilized at a value similar to E. coli ΔlonΔmgrB (Figure 3 B). Interestingly, a second GDA mutation (“GDA-2”, ∼2.2 Mb to 3.3 Mb) emerged in these populations (Figure 3 E, S4). The GDA-2 mutation was apparently lost and regained multiple times during the experiment (Figure 3 E, S4), suggesting a dynamic maintenance in the population rather than fixation. Further, the region of the genome amplified in GDA-2 was similar to what we had observed when E. coli Δlon had evolved in drug-free media (Figure S2, S4), indicating that GDA-2 was likely a fitness enhancing mutation unrelated to trimethoprim adaptation.

To further investigate the emergence of GDA-2 during evolution of LTMPR1 in drug-free media we sequenced 6 randomly picked colonies from each of the lineages at the final time point (Figure 3 F). All the sequenced isolates had a single copy of folA and a majority of isolates showed the presence of GDA-2 (Figure 3 F, Supplementary File 2). Interestingly, some of the isolates that did not harbour GDA-2 had evolved mutations in rpoS (Figure 3 F, Supplementary File 2). Loss-of-function mutations in RpoS are fixed at late time points during adaptation to trimethoprim and enhance the fitness of E. coli without conferring resistance [46, 51]. We also sequenced a few randomly picked isolates from WTMPR4 and WTMPR5 evolved lines (2 from each of the 3 replicate lineages in each case) to check for the presence of GDA-2 and rpoS mutations. As expected, clones derived from WTMPR4 had reverted to a single copy of folA, while WTMPR5 retained its ancestrally inherited folA mutation. None of them had evolved the GDA-2 mutation, while a few of the isolates from WTMPR4 and WTMPR5 lines had acquired rpoS mutations (Figure 3 G, H).

Consistent with earlier studies, and the role of mutations in rpoS in general fitness enhancement in E. coli, we found that an LTMPR1-derived isolate with mutant rpoS recovered fitness in drug-free media (Figure 4 A). Interestingly, an isolate with the GDA-2 mutation had also recovered fitness (Figure 4 A). Since GDA-2 occurred only in a lon-deficient background, and GDA-2 and rpoS mutations were mutually exclusive, we wondered whether the fitness enhancing effects of GDA-2 were also mediated by reduced RpoS activity. Lon is known to reduce the levels of RpoS during the exponential growth phase of E. coli [52]. Therefore, we reasoned that at least some targets of RpoS would be overexpressed in a lon-knockout. To test this prediction, we analysed the transcriptome of E. coli Δlon and specifically looked at the expression levels of 168 known targets of RpoS (Supplementary File 3). Of these, 41 genes were overexpressed in the lon-knockout by at least 1.5-fold over wild type, including genes induced in the stationary phase and by various stresses such as poxB, aidB and cbpA (Figure 4 B, C). Levels of rpoS mRNA were however unaffected in this strain. Next, we compared the expression levels of these 41 RpoS-targets in LTMPR1 and E. coli Δlon ΔmgrB with E. coli Δlon as the reference and found that most had similar expression levels in these three strains (Figure 4 C). Indeed a few of these genes showed higher expression levels in LTMPR1 and E. coli Δlon ΔmgrB, which was not unexpected since loss of mgrB is known to stabilize RpoS [53]. We checked their expression in isolates derived from LTMPR1 which harboured either GDA-2 or rpoS mutations. In both isolates RpoS target genes were down-regulated indicating a re-setting of RpoS activity (Figure 4 C). Further, in the isolate harbouring GDA-2, rpoS mRNA levels were themselves lowered to 0.49x of E. coli Δlon (Supplementary File 3). Finally, deletion of rpoS from E. coli Δlon enhanced the fitness of the strain (Figure 4 D). Thus, evolution of LTMPR1 in antibiotic-free media was characterized by rapid reversal of GDA-mediated resistance and acquisition of mutations that enhanced fitness by reducing the levels of RpoS-transcribed genes (Figure 4 E).

Lower RpoS activity leads to fitness enhancement of LTMPR1 in drug-free media.

A. Relative fitness (w) of isolates with the indicated genotypes derived from LTMPR1 evolution in antibiotic-free media (see also Figure 3) calculated using a competitive growth assay using E. coli ΔlonΔlacZ as the reference strain. Mean ± S.D. from three independent measurements is plotted. No change in relative fitness compared to the reference (w = 1) is shown as a dotted line for reference. Statistical significance was tested using an unpaired t- test. A p-value of < 0.05 was considered as a statistically significant difference (*). B. Expression level of 168 known targets of RpoS in E. coli Δlon compared to wild type expressed as log2(fold change) values determined by RNA sequencing. Dotted lines at 1.5- fold higher and lower than wild type represent cut-offs used to identify overexpressed and under-expressed genes. The pie chart shows the fraction of RpoS targets that were overexpressed in E. coli Δlon by at least 1.5-fold. C. Expression levels of 41 de-regulated RpoS targets in E. coli Δlon compared to wild type are shown as a heat map in the first vertical. The expression levels of these genes in LTMPR1, E. coli ΔlonΔmgrB or LTMPR1- derived isolates compared to E. coli Δlon are provided in the subsequent verticals. D. Relative fitness of E. coli ΔlonΔrpoS in antibiotic-free media calculated using a competitive growth assay using E. coli ΔlonΔlacZ as the reference strain. Neutrality of the ΔlacZ genetic marker was verified by competition between E. coli Δlon and E. coli ΔlonΔlacZ. Mean ± S.D. from three independent measurements is plotted. No change in relative fitness compared to the reference (w = 1) is shown as a dotted line for reference. Statistical significance was tested using an unpaired t-test. A p-value of < 0.5 was considered as a statistically significant difference (*) E. Model for the effects of Lon deficiency on RpoS levels and bacterial fitness. The roles of mutations within RpoS or the GDA-2 mutation in compensating for the defects of RpoS overproduction are shown.

Evolution at constant trimethoprim pressure

In the presence of trimethoprim (300 ng/mL), antibiotic resistance level and DHFR expression remained high over the course of evolution of LTMPR1 (Figure 5 A, B, C, D). WTMPR4 and WTMPR5 too remained resistant, and in the case of WTMPR4 an increased colony forming efficiency was observed at a late time point (Figure 5 C). These data suggested that under drug pressure high folA copy number was likely to be maintained. Indeed, population sequencing of LTMPR1 lineages showed that the ancestral GDA mutation was maintained until 196 generations of evolution in all three replicates (Figure 5 E, S5). By 252 generations, however, a majority of bacteria in Lines A and C had reverted to a single folA copy (Figure 5 E, S5) and had acquired a point mutation either in the folA gene (Line C) or its promoter (Lines A) (Figure 5 E, Supplementary File 4). Line B, on the other hand, retained high folA copy number, but with a reduction in the length of the genomic duplication (Figure 5 D, S5). Line B had also acquired a mutation in the folA promoter, however its frequency in the population was low (Figure 5 D, Supplementary File 4).

Evolutionary fate of folA-encompassing GDA in LTMPR1 at sustained trimethoprim pressure.

A. Schematic of the experimental pipeline used to investigate the impact of evolution in constant drug-pressure on resistance level, folA copy number, point mutations in folA and expression level of different trimethoprim resistant E. coli populations. B. Trimethoprim IC50 values of 3 evolving lineages starting from LTMPR1 (A, B, C) over 252 generations of evolution at 300 ng/mL trimethoprim. Mean ± S.D. from three measurements is plotted at each time point. IC50 values of E. coli Δlon and ΔlonΔmgrB are shown as dotted lines for reference. C. Colony formation of LTMPR1, WTMPR4 and WTMPR5 evolved in 300 ng/mL trimethoprim at indicated time points. Fraction of the population capable of forming colonies at 1, 5 and 10 μg/mL trimethoprim was calculated across 3 replicate lines. Mean value from the three lines is plotted. The results of similar experiments performed on ancestors (0 generations) are also provided for reference. D. DHFR expression in LTMPR1 lineages A, B and C evolved in 300 ng/mL trimethoprim, measured by immunoblotting using anti-DHFR polyclonal antibody. FtsZ was used as a loading control. Quantitation was performed by calculating band intensities using image analysis. DHFR expression at each time point was normalized to the ancestor (i.e. 0 generations, set to 1). Mean of three independent measurements is shown below each lane. E. Copy number of folA (GDA(folA)) at different time points of evolution of LTMPR1 in trimethoprim is plotted on the left y-axis. Copy number was determined by dividing number of reads from an Illumina sequencing experiment corresponding to folA by the average number of reads mapping to the rest of the genome. Frequency of various folA alleles in the evolving populations at each of the time points is plotted on the right y-axis. F. Point mutations in folA, rpoS and mgrB in 6 randomly picked trimethoprim-resistant colonies from each of the LTMPR1 lines at 252 generations of evolution are shown diagrammatically. Genotypes that were carried forward for further analysis are marked (* or #). G, H. Point mutations in folA, mgrB and rpoS in isolates derived from 252 generations of evolution of WTMPR4 (G) and WTMPR5 (H) in 300 ng/mL of trimethoprim. From each of the evolving lineages 2 random isolates were picked for genome sequencing. Various point mutants at the 3 gene loci are represented by appropriate symbols as shown in the legend.

In agreement with population sequencing, none of the resistant isolates derived from LTMPR1 Line A at the final time point showed the ancestral GDA. Instead, they all harboured a C-35T mutation in the folA promoter that is known to up-regulate DHFR expression [51] (Figure 5 F, Supplementary File 4). Among Line B isolates, many had retained the ancestral duplication, without acquiring any additional mutations in folA. One of the isolates from this Line that had reverted to a single folA copy had gained the A-64T folA promoter mutation (Figure 5 F, Supplementary File 4). In Line C too we found that isolates that had lost the ancestral GDA had evolved a point mutation in the folA promoter (+TG-30) or a Trp30Arg mutation in DHFR (Figure 5 F, Supplementary File 4). We found similar results in sequenced isolates from WTMPR4 lineages, i.e. loss of GDA and gain of a point mutation either in folA or its promoter (Figure 5 G). WTMPR5-derived isolates did not accumulate additional mutations in folA but did acquire mutations in rpoS (Figure 5 H). Interestingly, two isolates from LTMPR1 Lines B and C had evolved new GDA mutations encompassing folA. In one of the Line B isolates (Isolate 2) the ancestral GDA was replaced by 5x amplification of a shorter genomic region (∼12 kb) encompassing folA (Figure S6). Both ends of this newly amplified region were flanked by IS1 elements (Figure S6). On the other hand, one of the Line C isolates (Isolate 5) had an expanded duplication beyond what was originally found in LTMPR1 (Figure S6). Also, both copies of folA in this isolate coded for DHFR-Trp30Arg (Figure 5 E, S6), suggesting that the original GDA was lost and an extended GDA mutation was gained after fixation of the mutant folA allele.

Based on these results we concluded that even under drug pressure, GDA mutations were reversed in most bacteria. Notably, reversion to a single folA copy was slower than in drug-free media as it first required the fixation of point mutations at the folA locus. However, in a smaller sub-population, more than one folA gene copy was maintained. This maintenance was dynamic and likely involved repeated loss and gain of GDA mutations.

To further investigate the replacement of GDA by point mutants, we checked to see if point mutants could invade a population harbouring GDA mutations under drug pressure. We allowed LTMPR1-derived isolates with point mutations in folA (marked with ΔlacZ::Cat) to compete with a GDA-harbouring isolate from the same time point in evolution and asked how rapidly the former could invade a population of the latter. We performed these competitions with a contemporary GDA-harbouring isolate rather than the ancestor to nullify the effects of other media adaptations such as rpoS mutations that may have been acquired over evolution. Since for this assay we used point mutant strains genetically marked with ΔlacZ::Cat we verified the neutrality of the ΔlacZ::Cat marker by checking if the marked isolate could invade its unmarked parent. In the presence of trimethoprim (300 ng/mL) point mutants were enriched over the GDA mutant in co-culture by about 100-fold (Figure 6 A) recreating the results of our evolution experiment. A possible explanation for this finding may be higher resistance level conferred by folA point mutations than GDA. Indeed, some isolates with folA point mutations had higher IC50 values compared to a contemporary GDA-harbouring isolate (Figure 6 B). However, regardless of IC50 values, all isolates also had similar doubling times in the presence of 300 ng/mL trimethoprim (Figure 6 B). Thus, though there were no large observable differences in growth rate between isolates, point mutants could indeed invade and outcompete GDA mutations, reflecting either subtle differences in fitness between them or their relative stabilities in culture.

Invasion of GDA-dominant populations by point mutations under trimethoprim pressure.

A. Upper panel. Competition between isolates with point mutations in folA and an isolate that harboured a folA duplication derived from the same time point of LTMPR1 evolution (upper panel). The genotypes of the isolates (i.e. alleles at mgrB, folA and rpoS loci) used are shown diagrammatically (see also Figure 5, Supplementary File 4). The initial mixing ratio was 1000:1 in favour of the GDA mutant. Point mutants were marked genetically with ΔlacZ::Cat. Neutrality of the ΔlacZ::Cat marker was verified by competing marked and unmarked point mutants (lower panel). Percentage of the population constituted by the marked point mutants over ∼105 generations of serial transfer in media supplemented with trimethoprim (300 ng/mL) is plotted. Mean ± S.D. from 3 replicates are plotted and traces for individual mutants are appropriately colored. B. Trimethoprim IC50 values of the LTMPR1 ancestor or evolved isolates from 252 generations in trimethoprim with indicated genotypes (see also Figure 5, Supplementary File 4) are plotted as bars. Mean ± S.D. from three independent measurements are plotted. Statistical significance was tested using an unpaired t-test and LTMPR1 was used as the reference. A p-value < 0.05 was considered significantly different (*) while ≥ 0.5 was considered not statistically significant (ns). Doubling times relative to LTMPR1 (set to 1) in the presence of 300 ng/mL of trimethoprim are provided above the graph as mean ± S.D. from three independent measurements.

Maintenance of high folA gene copy number under drug pressure is driven by unstable mutants of DHFR

Though sustained drug pressure did extend the lifetime of folA gene duplication, acquisition of trimethoprim-resistant point mutations facilitated reversion to a single gene copy. We therefore asked whether folA duplication could be maintained under conditions of increasing antibiotic pressure, i.e., trimethoprim concentrations starting at 1 μg/mL and ending at 10 μg/mL over 252 generations of evolution (Figure 7 A). Under these conditions, trimethoprim IC50 values increased sharply over the course of evolution eventually reaching ∼40 to ∼160-fold higher than ancestral LTMPR1 (Figure 7 B). Colony formation at high trimethoprim concentrations also increased at later generations of evolution (Figure 7 C). Similarly evolved WTMPR4 and WTMPR5 also showed enhanced colony formation at high trimethoprim concentrations (Figure 7 C).

Evolutionary fate of folA-encompassing GDA in LTMPR1 at increasing trimethoprim pressure.

A. Schematic of the experimental pipeline used to investigate the impact of evolution in increasing drug-pressure on resistance level, folA copy number, point mutations in folA and expression level of different trimethoprim resistant E. coli populations. B. Trimethoprim IC50 values of 3 evolving lineages starting from LTMPR1 (A, B, C) over 252 generations of evolution are plotted on the left Y-axis. Mean ± S.D. from three measurements is plotted at each time point. Trimethoprim concentrations used during evolution are plotted in the right Y-axis. C. Colony formation of LTMPR1, WTMPR4 and WTMPR5 evolved in increasing trimethoprim at indicated time points. Fraction of the population capable of forming colonies at 1, 5 and 10 μg/mL trimethoprim was calculated across 3 replicate lines. Mean value from the three lines is plotted. The results of similar experiments performed on ancestors (0 generations) are also provided for reference. D. DHFR expression in LTMPR1 lineages A, B and C evolved in increasing trimethoprim, measured by immunoblotting using anti-DHFR polyclonal antibody. FtsZ was used as a loading control. Quantitation was performed by calculating band intensities using image analysis. DHFR expression at each time point was normalized to the ancestor (i.e. 0 generations, set to 1). Mean of three independent measurements is shown below each lane. E. Copy number of folA (GDA(folA)) at different time points of LTMPR1 evolution in increasing trimethoprim is plotted on the left y-axis. Copy number was determined by dividing number of reads from an Illumina sequencing experiment corresponding to folA by the average number of reads mapping to the rest of the genome. Frequency of various folA alleles in the evolving LTMPR1 populations at each of the time points is shown on the right y-axis. F. Point mutations in folA, rpoS and mgrB in 6 randomly-picked trimethoprim-resistant colonies from each of the LTMPR1 lines at 252 generations of evolution are shown schematically. G, H. Point mutations in folA, mgrB and rpoS in isolates derived from 252 generations of evolution of WTMPR4 (G) and WTMPR5 (H) in increasing trimethoprim. From each of the evolving lineages 2 random isolates were picked for genome sequencing. Various point mutants at the 3 gene loci are represented by appropriate symbols as shown in the legend.

Curiously, we observed that the 3 replicate lineages of LTMPR1 showed different trends in DHFR expression. Line A maintained a high level of DHFR protein for the duration of the experiment. On the other hand, Lines B and C showed high DHFR expression until 196 generations, after which it sharply declined (Figure 7 D). Genome sequencing revealed that all 3 lines convergently evolved the Trp30Arg mutation in DHFR early during evolution and had fixed this mutation by 196 generations. Until this time point, all 3 lines retained folA duplication (Figure 7 E, S7). Subsequently, Line A evolved an additional C-35T folA promoter mutation and had partially lost the ancestral GDA (Figure 7 E, S7, Supplementary File 5). In Lines B and C, a second missense mutation in DHFR, i.e. Tyr151Asp, appeared between 196 and 252 generations (Figure 7 D, Supplementary File 5). Line B also showed an Ile94Leu mutation in DHFR at low frequency at generation 252 (Figure 7 D, Supplementary File 5).

Unexpectedly, despite acquiring additional mutations and lower DHFR expression level, LTMPR1 Lines B and C retained two copies of folA in contrast to Line A (Figure 7 E, S7, Supplementary File 5).

Individual resistant clones from each of the lineages provided greater insight into the genetic trajectories of evolution. In LTMPR1 Line A, 2 of the 6 sequenced isolates had lost gene duplication and acquired the Trp30Arg coding and C-35T promoter mutations in folA (Figure 7 F, Supplementary File 5). Interestingly, the other 4 sequenced isolates from Line A retained the ancestral GDA and harboured the same combination of folA point mutations, but only in 1 of the 2 gene copies (Figure 7 F, Supplementary File 5). Ten of the twelve isolates from LTMPR1 Lines B and C, on the other hand, retained folA duplication. In all ten isolates, both copies of folA had acquired Trp30Arg and Tyr151Asp mutations (Figure 7 F, Supplementary File 5). Unlike LTMPR1, all sequenced isolates of WTMPR4 at 252 generations of evolution had lost their GDA mutation and gained 2 folA mutations (Figure 7 G). WTMPR5 too had gained an additional folA mutation (Figure 7 H). Notably, none of the isolates from WTMPR4 and WTMPR5 lineages had the Trp30Arg and Tyr151Asp combination of mutations that occurred repeatedly in LTMPR1 lineages.

We next sought to understand why some LTMPR1 bacteria had retained high folA copy number despite acquiring multiple point mutations in the drug target. To address this question, we first asked what the advantage of folA duplication was in the two different mutational trajectories. Isolates from Line A had similar IC50 values for trimethoprim regardless of whether they had one or two copies of folA (Figure 8 A). On the contrary, isolates from Line B that harboured only one copy of mutant folA (coding for DHFR-Trp30Arg-Tyr151Asp) displayed ∼3.5-fold lower IC50 than those that harboured two copies (Figure 8 A). Immunoblotting showed that isolates from Line A had similar expression level of DHFR (Figure 8 B). However, DHFR protein level was significantly compromised by the Trp30Arg-Tyr151Asp mutation combination and double the number of folA gene copies restored this defect (Figure 8 B). Thus, higher gene dosage of DHFR was dispensable once a promoter up-regulatory mutation was fixed. On the other hand, bacteria with alleles of DHFR harbouring multiple missense mutations continued to benefit from overexpression due to folA duplication.

Proteostatic pressure facilitates maintenance of folA duplication

A. Trimethoprim-IC50 values of resistant isolates with the indicated genotypes (see also Figure 7) derived from LTMPR1 evolution in increasing antibiotic pressure. Mean ± S.D. from three independent measurements is plotted. Statistical significance was tested using an unpaired t-test. A p-value of <0.05 was considered statistically significant (*). A p-value > 0.05 was considered not statistically significant (ns). B. DHFR expression level in LTMPR1-derived trimethoprim resistant isolates with indicated genotypes assessed by immunoblotting using anti-DHFR polyclonal antibody. FtsZ was used as a loading control. Quantitation was performed by calculating band intensities using image analysis. DHFR expression was normalized to the ancestor (LTMPR1). Mean of three independent measurements is provided. C. Cartoon representation of the structure of E. coli DHFR (PDB: 7DFR) bound to folate. Residues Trp30, Tyr151, Phe153 and Ile155 which form hydrophobic interactions and are required for proteolytic stability are shown as sticks and coloured by element (C: green, O:red, N:blue). Distances of less than or equal to 4 Å, which indicates possible interactions, are shown as dotted yellow traces. Folate bound in the active site of DHFR is shown as sticks. D. Schematic of the experiment used to test trimethoprim resistance and expression level of various DHFR mutants in E. coli wild type (WT) or Δlon. E. Trimethoprim IC50 values of E. coli wild type or E. coli Δlon heterologously expressing DHFR (wt) or its mutants. Mean ± S.D. from three independent measurements is plotted. Statistical significance was tested using an unpaired t-test. Comparisons were between mutant and wild type DHFR, unless otherwise indicated. A p-value of <0.05 was considered statistically significant (*). A p-value > 0.05 was considered not statistically significant (ns). F. Expression level of plasmid-borne DHFR (wt) or its mutants in wild type or Δlon E. coli assessed by immunoblotting using anti-DHFR polyclonal antibody. FtsZ was used as a loading control. Quantitation was performed by calculating band intensities using image analysis. Expression of mutants was normalized to wild type DHFR (set to 1). Mean of three independent measurements is shown.

Mutations at position Trp30 in DHFR render the protein resistant to inhibition by trimethoprim but are accompanied by lower in vivo proteolytic stability due to perturbation of the hydrophobic core of the protein [54]. This affects the expression level of DHFR, resulting in lower cellular abundance that limits the resistance level conferred by mutations at Trp30 [49, 54]. Interestingly, residue Tyr151 lies in the same hydrophobic pocket as Trp30 (Figure 8 C). To test the effect of combination of mutations at these sites on trimethoprim resistance and DHFR expression, we first heterologously expressed wild type DHFR, Trp30Arg or Tyr151Asp single mutants and the Trp30Arg-Tyr151Asp double mutant in wild type E. coli (Figure 8 D). As expected, DHFR Trp30Arg expression increased the IC50 of trimethoprim. Surprisingly, neither DHFR Tyr151Asp, nor the Trp30Arg-Tyr151Asp double mutant conferred resistance to the antibiotic (Figure 8 E). Indeed, both mutants conferred lower IC50 values than wild type DHFR. This effect was due to loss of hydrophobicity at the Tyr151 position, as DHFR-Tyr151Leu and DHFR-Tyr151Phe conferred similar IC50 values as wild type DHFR (Figure 8 E). Since the Trp30Arg-Tyr151Asp mutation combination had evolved in a lon-deficient background, we next expressed these mutants in E. coli Δlon (Figure 8 D). Here too, DHFR Trp30Arg conferred trimethoprim resistance, while the Tyr151Asp mutant did not (Figure 8 E). Strikingly, the combination of Trp30Arg and Tyr151Asp conferred very high-level trimethoprim resistance (Figure 8 E) demonstrating synergistic epistasis that was contingent on the absence of Lon protease. The levels of expression of DHFR mutants in wild type and lon-deficient bacteria mirrored trimethoprim resistance levels indicating that clearance by Lon altered their genotype-phenotype relationship (Figure 8 F). Based on these data, we concluded that the action of proteostatic machinery generated additional demand for expression by clearing mutant-DHFR proteins. This additional demand drove the maintenance of folA duplications in bacteria evolving at very high drug pressures.

Discussion

Ohno’s hypothesis, proposed in 1970, provided a conceptual framework to think about the evolution and eventual fate of duplicated genes [55]. Several theoretical and experimental studies have built upon the original idea to delineate the evolutionary outcomes of gene duplication [18, 56, 57]. More recently, experimental evolution of microbes has been used to test some of the suggestions made by Ohno and others. This approach has the inherent advantage of greater control over the selection pressure applied and availability of genetic tools to interrogate molecular mechanisms. In most studies, an artificial system was constructed and then evolved under defined selection pressures [33, 58–61]. Our study began with serendipitous isolation of spontaneous trimethoprim-resistant mutants of E. coli that overproduced DHFR because of a folA gene duplication. We exploited this system to understand the relationship between selection pressure and gene duplication in the context of drug-resistance evolution. The selection pressure for higher gene expression, i.e. “expression demand”, can be understood as the amount of a gene product that is required to maintain fitness in a particular environment. Antibiotics directly increase expression demand by inhibiting the activity of their targets. Our study shows that increasing gene dosage is a rapid, but transient evolutionary response to ensure that expression demand is met. On the other hand, point mutations are relatively stable routes to mitigating expression demand. Over short time scales, changes in gene dosage can counter the effects of antibiotic. However, their inherent instability results in almost inevitable replacement with phenotypically equivalent point mutations. Thus, continued antibiotic exposure delayed but could not prevent reversion to low gene copy number in most of our experiments. Further, we discover a novel association between proteostasis and gene dosage evolution. In E. coli evolving at high trimethoprim pressure, bacterial protein quality control exacerbated expression demand by clearing mutant DHFR-s. As a result, drug-resistant mutants of DHFR with low proteolytic stability drove the maintenance of folA gene duplication (Figure 9). Based on these findings we propose that a combination of factors, which include environmental and physiological, alter the relationship between gene copy number and fitness, and drive gene dosage evolution.

Model of expression demand induced selection of folA gene duplication, followed by replacement by phenotypically equivalent point mutants. Expression demand for DHFR is generated by trimethoprim pressure, which results in the selection of folA gene duplications. Gene duplication is inherently unstable and reverses to a single copy of wild type folA when drug-pressure is withdrawn. Under drug pressure, folA gene duplication is maintained until a point mutation that confers resistance arises in one of the copies of folA (promoter or coding region). Upon acquisition of a point mutation, reversal of gene duplication ensues unless additional expression demand is generated by the action of proteostatic machinery on unstable drug-resistant DHFR mutants.

A key finding from our study was the two-fold influence that proteostatic machinery had on gene copy number. Firstly, Lon protease-deficiency promoted GDA mutations, possibly by allowing accumulation of IS-tranposase proteins in E. coli. This observation resonates with earlier work by Nicoloff and co-workers who found that large IS-element flanked genomic duplications in tetracycline-resistant E. coli were commonly preceded by spontaneous mutations that inactivated lon [62, 63]. Further, the role of Lon protease in regulating IS-element mobility by a similar mechanism has been previously demonstrated for IS1 and IS903 [64, 65]. We note however that other mechanisms by which Lon may modulate the frequency of IS-mediated GDAs in E. coli cannot be ruled out. For instance, Lon protease homologs from several organisms directly bind DNA [66–70], though the role of this binding in DNA transactions is unclear. Similarly, Lon is a regulator of the recombination protein, RecA, by its action on LexA [71] and in Salmonella RecA stimulated the rate of gene duplications mediated by IS3 [23]. The second impact of protein quality control on GDA was the action of proteases on trimethoprim-resistant mutants of DHFR, which promoted gene duplication by generating additional demand for expression (Figure 9). Importantly, which of these two effects dominated was determined by a combination of environment, i.e. drug concentrations, and other mutations acquired at the folA locus during evolution. In particular, mutations in the folA promoter could alleviate expression demand and hence reduce selection for high gene copy number.

The interplay between GDA and point mutations during evolution that we report here agrees in many respects with a recent study by Tomanek and Guet [61]. They use an elegantly designed artificial system [33, 61] in which expression of the galK gene in E. coli was driven by a very low activity promoter. This strain was then allowed to evolve in different concentrations of galactose. Like the present study, they found that the activity of IS-elements biased evolutionary trajectories in favor of duplications. They also showed that co-occurrence of duplications and point mutations was selected only at high expression demand, in agreement with our findings. Finally, they argued that gene duplication events may constrain evolution by point mutations. Our study corroborates these results at the short term in that the lon-knockout which showed a higher propensity for GDA mutations did not evolve point mutations in folA at early time points of evolution. However, given the relative low stability of large GDAs, these were inevitably lost over longer duration of evolution and replaced with point mutations. Thus, it is likely that the different mutational profiles of GDA-biased populations may be relevant only at short durations of evolution. It is important to note that though we have focused on the effects of mutations at the folA locus, which are directly under trimethoprim pressure, many of the isolates from our evolution experiments also harboured additional mutations at other loci. For instance, mutations in rpoS were seen throughout our evolution lines and at least in drug-free media were significant contributors to bacterial fitness. It is not yet clear what the role of other mutations are likely to be under antibiotic selection, though it may be envisaged that these additional mutations play the role of media adaptations. We did not take their effects into in this study account since we expected them to be minor contributors to fitness especially during evolution under trimethoprim selection. However, their role in modulating the stability and selection of GDA mutations in drug-free/low drug environments may be more significant and remains to be investigated in the future.

In summary, the present study uncovers the complex relationship between gene copy number, point mutations and environment, and identifies proteostasis as a regulator of gene dosage evolution in bacteria. These findings open up the possibility of alternative explanations for how gene duplications may be maintained over long periods of evolution and provides insight into how gene expression demand is mitigated by multiple mechanisms in bacteria. Moreover, our findings relating proteostasis in general and Lon in particular, to adaptive mutations add to a growing body of literature that define a causal link between protein quality control and mutational paths of evolution. For DHFR itself, decreasing the stringency of protein quality control changes the fitness landscape of single mutations and their combinations, altering the mutational repertoires accessible during adaptive evolution [49, 50, 72, 73]. Similar effects of protein homeostatic machinery on mutant fitness are reported from a diverse range of proteins that influence phenomena like viral evolution [74] and the fate of horizontally acquired genes [75–77]. Indeed, a central role for protein quality control machinery in molecular and organismal evolution is reflected in conserved signatures of the proteostatic network across the different kingdoms of life [78]. These links may be important to consider for a new wave of potential antibacterials that target proteostasis [79–81]. It appears critical to ensure that the impact of changes in evolvability are considered when designing such strategies for therapeutic intervention.

Materials and methods

Strains, plasmids and culture conditions

Escherichia coli K-12 MG1655 was used as wild type for all experiments. E. coli DH10B was used for plasmid construction and manipulation. E. coli strains were cultured in Luria-Bertani (LB) broth with shaking at 180-200 rpm or on LB agar plates at 37 °C. Ampicillin was used at a concentration of 100 μg/mL for selection of plasmid-transformed strains. Trimethoprim was added to growth media at required concentrations as needed. The strains and plasmids used in this study are listed in Table 2.

Laboratory evolution of antibiotic resistance

Short-term evolution of antibiotic resistance

For short term laboratory evolution of antibiotic resistance, wild type and E. coli Δlon populations were cultured in 96-well plates in LB medium incorporating sub-inhibitory concentrations of different antibiotics. Following concentrations of antibiotics were used: trimethoprim (300 ng/mL; MIC/3); spectinomycin (5 μg/mL; MIC/10); nalidixic acid (4.25 μg/mL; MIC/3); erythromycin (10 μg/mL; MIC/10); rifampicin (4.25 μg/mL; MIC/3). Control populations were also set up in antibiotic-free LB. Five replicates were included for each strain in each antibiotic. Total volume of culture in each well was 150 μL. 10 % of the population was passaged to fresh growth media every 24 hours, for 7 days, resulting in ∼25 generations of evolution. At the end of 7 days, the evolved populations were frozen as glycerol stocks.

To check for the evolution of resistance and isolation of resistant clones for further analyses, frozen populations were revived in drug-free LB broth and grown overnight (14-16 hours). Saturated cultures were serially diluted and spotted on LB agar plates supplemented with the appropriate antibiotic at MIC. Colonies were randomly picked, cultured overnight in drug-free LB and frozen in 50% glycerol at-80 °C until further use.

Long-term evolution at different trimethoprim pressures

For long-term evolution, trimethoprim-resistant isolates LTMPR1, WTMPR4 and WTMPR5 were first revived from frozen stocks in drug-free LB overnight. Triplicate evolving lineages were then established by sub-culturing 1% of the saturated culture of each ancestor into 2 mL of appropriate medium. For relaxed selection drug-free LB was used. For constant antibiotic pressure, trimethoprim was added at a concentration of 300 ng/mL. For increasing antibiotic pressure, trimethoprim was increased from 1 μg/mL to 10 μg/mL over course of the experiment as shown in Figure 7 B. Cultures were grown for 10-14 hours (∼7 generations) and then passaged (1%) into fresh media. This was continued for 36 passages, i.e. ∼252 generations. Aliquots of cultures were frozen periodically for further analyses.

For isolating individual clones, populations were revived in drug-free LB from frozen stocks at 252 generations of evolution in drug-free LB, serially diluted and then spotted onto LB agar plates supplemented with trimethoprim (1, 5 or 10 µg/mL). Plates were incubated for 18-20 hours at 37 °C and resistant colonies were picked randomly for further analyses.

Measuring trimethoprim resistance

Trimethoprim IC50

Trimethoprim Inhibitory Concentration - 50 (IC50), i.e. the antibiotic concentration needed for 50 % growth inhibition, was calculated using a broth dilution assay. Frozen glycerol stocks of appropriate bacterial strains were revived in 1-3 mL LB for 6-8 hours. In a 96-well plate, 150 μL of LB was added to individual wells. Trimethoprim stocks were added, and the antibiotic was serially diluted. No antibiotic was added to a few wells, which served as controls. Next, 1.5 μL of the appropriate bacterial culture was added to each well. The peripheral wells of the plate were filled with water to prevent dehydration. The 96-well plate was then incubated at 37LJ for 15-18 hrs. Optical density at 600 nm was measured for each well using a plate reader (Ensight, Perkin Elmer). Obtained data were first normalised to growth in drug-free wells (set to 1) and then fitted to a sigmoidal log(inhibitor)-response curve using Graphpad Prism software. The equation for this model is as follows:

Y= Min + (Max-Min)/(1+10 (X-Log(IC50)))

where, Min: Minimum value of normalised OD and Max: maximum value of normalised OD Values of IC50 were estimated from the above curve fit and averaged across 3 independent measurements for every strain.

Colony forming efficiency

To calculate the colony forming efficiency of evolving populations at different trimethoprim concentrations, frozen glycerol stocks were first revived in 1 mL LB for 14-15 hours. Saturated cultures were serially diluted (10-fold). Diluted cultures were then spotted (10 mL) on LB agar plates containing 0, 1, 5 and 10 μg/mL of trimethoprim. Spots were allowed to dry completely and the plates were incubated at 37 °C for ∼18 hours. Colonies were counted from the least dense spot and multiplied by the appropriate dilution factor to estimate colony forming units per mL (CFUs/mL) of culture.

Growth curves

Appropriate frozen stocks were revived overnight in drug-free LB (1-3 mL) and 1μl of the saturated culture was added to 200 μL of media supplemented with trimethoprim (300 ng/ml) in the wells of a sterile uncoated 96-well plate. Optical density at 600 nm was measured during growth at 37 °C at 15-minute intervals over a 24-hour period on a Perkin-Elmer Ensight Multimode Plate Reader. Doubling time (DT) (in mins) was estimated from log phase, using the following formula: Where OD1 and OD2 represent optical densities at the onset and end of log phase respectively, and T1 and T2 represent time elapsed (in mins), corresponding to OD1 and OD2.

Genome sequencing, identification of mutations and detection of GDA mutations

Genomic DNA was extracted from late log-phase cultures of appropriate strain/resistant isolates/evolving populations using phenol:chloroform:isoamyl alcohol (25:24:1) extraction as previously described [82]. Extracted genomic DNA was then cleaned-up using HiGenoMB DNA isolation and purification kit (Himedia, India) as per manufacturer’s instructions and quantitated spectrophotometrically. The integrity was checked by running purified DNA on an agarose gel before further analyses.

For whole-genome resequencing, paired-end whole-genome next-generation sequencing (NGS) was performed on a MiSeq system (Illumina, USA) with read lengths of 150-250 bps. Library preparation and sequencing services were provided by Eurofins (Bangalore, India) and Genotypic (Bangalore, India). Processed reads were aligned to the reference genome E. coli K-12 MG1655 (U00096) using bowtie2. All sequenced genomes had an average coverage depth of at least 100x. Variant calling and prediction of new junctions in the genome to identify large structural mutations was done using Breseq [83] using default settings in population or clonal modes as needed. All variants and new junctions that were already present in the ancestor were excluded. Variants unique to the test sample were classified as “mutations”.

For detection of GDA mutations, ‘coverage’ vs ‘coordinate in reference genome’ plot generated by the bam.cov command in Breseq was used [83]. Contiguous stretches greater than 10 kb with coverage depth of 1.5x or more, compared to the rest of the genome, were classified as GDA mutations. The presence of IS-elements or repeat sequences was manually checked using the coordinates of the GDA.

Calculating relative fitness of E. coli strains

Relative fitness of gene knockouts or evolved clones was estimated using competition assays. Saturated cultures of test strains and reference strains (E. coli ΔlacZ or E. coli ΔlacZΔlon) were mixed 1:1 to a total of 1% in 3 mL LB. Strains were allowed to compete for ∼7 generations (24 hours incubation). Mixed cultures were diluted and plated on LB agar supplemented with IPTG (1 mM) and X-gal (50 μg/mL) at the start and end of the competition experiment. Plates were incubated for 18-24 hrs, and blue and white colonies were counted to determine the CFUs/mL at initial and final time points.

Relative fitness (w) was calculated using the following formula: where Tf and Ti are the final and initial CFUs/mL of the test strain and Rf and Ri are the final and initial CFUs/mL of the reference strain, respectively.

Estimation of DHFR expression by immunoblotting

Immunoblotting was used to determine the expression level of DHFR in various strains/mutants of E. coli as described in Patel and Matange (2021) [46]. Briefly, equal lysate protein (5 μg) was subjected to SDS-PAGE on a 15% gel and electroblotted onto PVDF membranes. The membrane was blocked with 5% BSA solution and cut into 2 halves at the 25 kDa molecular weight band. The lower part of the membrane was probed with anti-DHFR polyclonal IgG (100 ng/mL) and the upper part was probed with anti-FtsZ serum (1:50,000). HRP-linked anti-rabbit IgG (1:10,000) was used as the secondary antibody. Hybridised antibodies were detected using chemiluminescence. Band intensities were quantitated in ImageJ and normalised to appropriate control (set to 1). Fold change in DHFR was calculated as follows:

Gene expression analysis using RNA-sequencing

Transcriptome analyses were performed using RNA sequencing. For RNA-seq, saturated overnight culture of appropriate strains/isolates was inoculated (1 %) into 3 mL LB in triplicate and incubated at 37 °C for 3 hours with shaking at 180 rpm. The cultures were centrifuged and bacterial pellets were resuspended in 1 mL of RNAlater (Invitrogen) for storage. RNA extraction, sequencing and preliminary data analyses were performed by Eurofins (Bangalore, India). Total RNA was extracted using RNAeasy spin columns and quantitated using Qubit and Bioanalyzer. Bacterial rRNA was depleted using Ribozero kit. Bacterial mRNA was then reverse transcribed, library prepared and quality control performed using Tapestation platform. Paired end sequencing was performed on Illumina platform with read lengths of 150 bp read length. Processed reads were aligned to the reference genome (U00096.3) using HISTAT2(version 2.1.0). Abundance estimation was done using featureCounts(version1.34.0). Differential gene expression (DGE) analysis was done by comparing the expression of individual genes in test and reference samples. The output of the DGE analyses were log2(fold change) values for each gene and P-values for statistical significance.

Cloning and site-directed mutagenesis

Cloning of InsB and InsL transposases

InsB was amplified from E. coli K-12 MG1655 genomic DNA by PCR using primers insB1_fwd_NcoI (5’-GGCGCCATGGATATGCCGGGCAACAGCCCG-3’) insB1_rev_EcoRI (5’- CCTTTGAATTCTTATTGATAGTGTTTTATG-3’). For amplification of insL, primers insL_EcoRI_fwd (5’- ATGCCGAATTCATGAATTACTCTCACGATAAC- 3’) and insL_HindIII_rev (5’- ATGCCAAGCTTTTAGTTCTTCTTTTCGGATCC-3’) were used. Amplified insB gene was digested with NcoI and EcoRI restriction endonuclease and ligated with similarly digested pPROEx-HtC plasmid. Amplified insL was digested with EcoRI and HindIII restriction endonucleases and ligated with similarly digested pPROEX- HTa plasmid. Positive clones were screening using restriction digestion. Clones were transformed into E. coli wild type or Δlon for further analyses.

Site-directed mutagenesis for DHFR

Site-directed mutagenesis was performed to introduce mutations into the coding sequence of folA in pPRO-folA as described in [54]. Briefly, PCR was performed using pPRO-folA or pPRO-folA Trp30Arg as template and the appropriate forward and reverse mutagenic primers given below:

folA_Tyr151Asp_fwd_BglII:

5’-CTCTCACAGCGATTGCTTTGAGATCTTGGAGCGGCGGGGCC-3’

folA_Tyr151Asp_rev_BglII:

5’-CGCCGCTCCAAGATCTCAAAGCAATCGCTGTGAGAGTTCTGCG-3’

folA_Tyr151Leu_fwd_BglII:

5’-TCTCACAGCTTATGCTTTGAGATCTTGGAGCGGCGGGGCC-3’

folA_Tyr151Leu_rev_BglII:

5’- CGCCGCTCCAAGATCTCAAAGCATAAGCTGTGAGAGTTCTGCG-3’

folA_Tyr151Phe_fwd_BglII:

5’- CTCACAGCTTTTGCTTTGAGATCTTGGAGCGGCGGGGCC-3’

folA_Tyr151Phe_rev_BglII:

5’- GCCGCTCCAAGATCTCAAAGCAAAAGCTGTGAGAGTTCTGCG-3’

PCR products were digested with DpnI overnight, transformed into chemically competent E. coli DH10B and selected on LB agar plates containing ampicillin. Plasmids habouring the desired mutations were screened by restriction digestion and confirmed by sequencing. Sequenced clones were transformed into E. coli wild type or Δlon for further analyses.

In-cell stability of InsB and InsL

A chloramphenicol-chase assay was used to assess the in-cell stability of heterologously expressed InsB and InsL transposases as described in Matange (2020) [49] with a few modifications. Overnight cultures of E. coli wild type or Δlon harbouring plasmids pPRO- insB or pPRO-insL were diluted 1% in 5 mL LB supplemented with ampicillin and IPTG (0.1mM). After 4 hours of growth at 37 °C, 1 mL of culture was pelleted. The cell pellet was lysed in 100 μL of 1x Laemmli sample loading buffer by heating at 95°C for 10-20 mins. Chloramphenicol was added to the remaining culture at a final concentration of 25 μg/mL to stop protein synthesis and the culture was placed back in the incubator. After 15, 45, and 90 minutes 1 mL aliquots of the culture were pelleted and lysed. Lysates were subjected to SDS- PAGE followed by immunoblotting to detect the levels of InsB and InsL. Anti-Hexahis monoclonal antibody (Invitrogen) was used as the primary antibody for InsL. Anti-Hexahis polyclonal antibody (Invitrogen) was used as the primary antibody for InsB. Band intensities were calculated in ImageJ and normalised to control (set to 1) to calculate fraction of protein remaining after chloramphenicol treatment.